### 2013-09-15 project001/UVM/paper/mcrn501-p1-uvm.abs.tex
working on our paper, integrating paillard model, downloading more data

The explanatory value of simple climate models relies upon their ability to mechanistically explain past variation while providing a reasonable fit to available data.
We investigate the proper fitting of Paillard's three state model with past data to test its accuracy directly.
The parameter fit, our main effort, is achieved via a covariance matrix adaptation evolution strategy (CMA-ES) and a general genetic algorithm (GA).
We find that Paillard's model's explanatory value is mainly empirical, as the model is robust to parameter changes and the fit to data from $\delta ^{18} O$ records has minimum RMSE 1e-5.

### 2013-09-15 project001/UVM/paper/mcrn501-p1-uvm.body.tex
working on our paper, integrating paillard model, downloading more data

\section{Introduction}

Understanding the past climate of the earth allows for us to begin to understand what effects we are having upon it, and to predict what the future will bring.
From ice cores, a rough temperature record of Earth is constructed and we explain attempt to explain that.
The variation in the orbital parameters of our planet affect the amount of solar insolation recieved, and the timescale variation of these quantities is one possible forcing for past climatic change.
The eliptical elements of the orbit are changing slowly in time, as is the orientation of the planet's spin axis.
A three state model from Paillard fits the data emprically, and we attempt a rigorous fit of this model to data using the tools of evolutionary computation.

\section{Paillard's Model}

With some difficulty understanding his original paper \cite{paillard1998timing}, we recontruct Paillard's model for both versions that he provides.
The following two subsections paraphrase \cite{paillard1998timing} directly.

\subsection{Discrete model}

\newcommand{\inter}{\textbf{i}}
\newcommand{\mild}{\textbf{g}}
\newcommand{\full}{\textbf{G}}

The model has three states called \inter~(interglacial), \mild~(mild glacial) and \full~(full glacial).
This model undergoes an \inter-\mild~transistion as soon as the insolation falls below $i_0$.
An \mild-\full~transistion happens when the ice volume exceeds a threshold $v_\text{max}$ and finally a transistion \full-\inter~occurs when the insolation increases above $i_1$.
These transistions are assumed to be one-way, and the only ones allowed are those just discussed.
The \mild-\full~transition has the constraints that it will happen after some time $t_g$ and with insolation lower than some level $i_3$.
If $t_g$ is exceeded but $i_3$ is also exceeded, then the \mild-\full~transistion occurs as the next insolation decrease: when insolation falls below $i_2$.
% THIS MODEL IS WEIRDDDDDD

The part which was made least clear was whether the insolation exceeding $i_3$ necessitates a reset of the time in state \mild until $t_g$ is reached.
To achieve similar results, we assume that the answer is yes, the time does reset.

\subsection{Differential equation model}

This model is an extension of the discrete model.
There are still three regimes but now we define the ice volume $v$ as a differential equation

$$ \frac{dv}{dt} = (v_R - v) /\tau _R - F/\tau _F .$$

where $R$ is the current climate regime ($R$ = \inter,\mild,\full), $v_R$ are the reference ice volumes for the different regimes, $F$ is the forcing and $\tau _{R,F}$ are the time constants.
The transistions are the same for \inter-\mild~and \full-\mild~but \mild-\full~now happens when $v \geq v_\text{max}$, some threshold value (without any condition on the insolation).
Paillard also ``normalizes the ice volume to unity'': $v_g = v_G = v_\text{max} = 1$ and $v_i = 0$.
For the finale of model weirdness, the forcing $F$ is defined as a truncated insolation signal with truncation function $f$ as

$$ f(x) = \frac{1}{2} \left ( x + \sqrt{4a^2 + x^2} \right ) $$

and setting $a=1$.

\section{Computing Insolation}

The only input to the model is the solar insolation, so we are particulary careful to obtain accurate values.
Increasingly accurate integrations of the state of the solar system have been performed dating from the time of Lagrange in 1781 through Milankovich's use of Pilgrim's 1904 computations to establish his theory of past climate change to today's direct numerical integrations \cite{laskar2004long}.

Using the orbital parameters of eccentricity $e$, obliquity $\epsilon$, and longitude of the parahelion $\overline{\omega}$ computed from an integration function insola.f provided from Laskar \cite{laskar2004long} we can reconstruct the insolation at any time.
The following derivation is due to Berger \cite{berger1981long}.

The insolation $Q$ at a given location and time can be written as a function of the solar constant $S_0$, the distance from the Earth to sun relative to the average $R_0/R_E$ the solar zenith angle $\Theta$:

$$ Q = S_0 \frac{R_0^2}{R_E^2} \cos (\Theta) .$$

To solve for the average insolation during one day at a given latitute $\phi$, we need to express the solar zenith angle $\Theta$ in terms of $\phi$ and the known orbital parameters.
Taking $\delta$ as the latitude of the point directly subsolar and $h$ as the hour angle, we use the spherical law of cosines to write

$$ \cos ( \Theta ) = \sin ( \phi ) \sin ( \delta ) + \cos ( \phi ) \cos ( \delta ) \cos ( h ) .$$

Now to have the average daily insolation $\overline{Q} _\text{day}$ we take the integral of $Q$ from $h_o $ to $-h_0$ where $h_0$ is the hour angle for which the sun first rises.
Since we are concerned with the $\phi = $65N, the sun always rises and sets so we won't consider the cases for the sun neither rising nor setting.
Integrating, we have

$$ \overline{Q} _\text{day} = \frac{S_0}{\pi } \frac{R_0^2}{R_E^2} \left [ h_0 \sin (\phi) \sin (\delta) + \cos (\phi) \cos (\delta) \sin (h_0) \right ] . $$

Defining $\theta = 0$ the be the vernal equinox, we can write $\delta$ and $R_0/R_E$ as a function of the orbital parameters in a simple form at the summer solsctice ($\theta = \pi /2$):

$$ \delta = \epsilon ~~~~\text{and} ~~~~ \frac{R_0}{R_E} = 1 + e\sin (\overline {\omega}) .$$

\section{Evolutionary Algorithm Methods}

Morgan

\section{Results}

Figure \ref{fig:insol-data} depicts the reconstruction of insolation data back in time.

\begin{figure}[tpb!]
\centering
  \includegraphics[width=.48\textwidth]{../data/insolation/insol_data2_noname.pdf}
  \caption{
    The average daily solar insolation $\overline{Q}_{\text{day}}$ is computed on the summer solstice at 65N with the prodived $e,\epsilon,$ and $\overline{\omega}$.
  }
  \label{fig:insol-data}
\end{figure}

Figure \ref{fig:paillard-fig3} is a direct representation of the second figure of Paillard's original paper \cite{paillard1998timing}.

\begin{figure}[tpb!]
\centering
  \includegraphics[width=.53\textwidth]{paillardFig3.pdf}
  \caption{
    A first run of the Paillard ODE model, over the $\delta ^{18} O$ record show a reasonable emprical fit.
  }
  \label{fig:paillard-fig3}
\end{figure}

\begin{figure}[tpb!]
\centering
  \includegraphics[width=.45\textwidth]{../src/discrete_plot_noname.pdf}
  \caption{
    The discrete Paillard model.
  }
  \label{fig:paillard-fig3}
\end{figure}

\section{Conclusion}

The fitted Paillard-Reagan-Frank model of climatic variability wholly explains the temperature variation in the $\delta ^{18} O$ record.
With only forcing from changing solar insolation, this verifies that past climatic varaition of larger time scales is a result of the variation of the Earth's orbital parameters.

### 2013-09-15 project001/UVM/data/insolation/laskar2004/plot_data.m

@@ -1,0 +1,182 @@
+% % insolation data calculation
+% 
+% % http://en.wikipedia.org/wiki/Milankovitch_cycles
+% 
+% load INSOLN.LA2004.BTL.ASC
+% 
+% plot_back = 800;
+% 
+% close all;
+% figure;
+% subplot(6,1,2);
+% plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,2));
+% title('eccentricity e')
+% 
+% subplot(6,1,1);
+% plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,3));
+% title('obliquity \epsilon')
+% 
+% subplot(6,1,3);
+% plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),sin(INSOLN_LA2004_BTL_ASC(1:plot_back,4)));
+% title('sin(\omega)')
+% 
+% subplot(6,1,4);
+% plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,2).*sin(INSOLN_LA2004_BTL_ASC(1:plot_back,4)));
+% title('e sin(\omega)')
+% 
+% s0 = 1367; % solar constant
+% phi = deg2rad(65); % latitude
+% theta = pi/2; % summer solstice
+% 
+% % insolation = zeros(length(INSOLN_LA2004_BTL_ASC(:,1)),1);
+% % 
+% % for i=1:length(INSOLN_LA2004_BTL_ASC(:,1))
+% % 
+% % e = INSOLN_LA2004_BTL_ASC(i,2); % eccentricity
+% % eps = INSOLN_LA2004_BTL_ASC(i,3); % obliquity
+% % pibar = INSOLN_LA2004_BTL_ASC(i,4); % perihelion distance
+% % 
+% % dist_sun = (1+e*cos(theta - pibar)); %/(1-e^2);
+% % 
+% % delta = eps; % at the summer solstice
+% % 
+% % h0 = acos(-tan(phi)*tan(delta));
+% % q = s0/pi*dist_sun^2*(h0*sin(phi)*sin(delta)+cos(phi)*cos(delta)*sin(h0));
+% % 
+% % insolation(i) = q;
+% % end
+% 
+% subplot(6,1,5);
[…]

### 2013-09-15 project001/UVM/data/insolation/plot_data.m

@@ -1,59 +1,0 @@
-% insolation data calculation
-
-% http://en.wikipedia.org/wiki/Milankovitch_cycles
-
-load INSOLN.LA2004.BTL.ASC
-
-plot_back = 800;
-
-close all;
-figure;
-subplot(6,1,2);
-plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,2));
-title('eccentricity e')
-
-subplot(6,1,1);
-plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,3));
-title('obliquity \epsilon')
-
-subplot(6,1,3);
-plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),sin(INSOLN_LA2004_BTL_ASC(1:plot_back,4)));
-title('sin(\omega)')
-
-subplot(6,1,4);
-plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),INSOLN_LA2004_BTL_ASC(1:plot_back,2).*sin(INSOLN_LA2004_BTL_ASC(1:plot_back,4)));
-title('e sin(\omega)')
-
-s0 = 1367; % solar constant
-phi = deg2rad(65); % latitude
-theta = pi/2; % summer solstice
-
-insolation = zeros(length(INSOLN_LA2004_BTL_ASC(:,1)),1);
-
-for i=1:length(INSOLN_LA2004_BTL_ASC(:,1))
-
-e = INSOLN_LA2004_BTL_ASC(i,2); % eccentricity
-eps = INSOLN_LA2004_BTL_ASC(i,3); % obliquity
-pibar = INSOLN_LA2004_BTL_ASC(i,4); % perihelion distance
-
-dist_sun = (1+e*cos(theta - pibar)); %/(1-e^2);
-
-delta = eps; % at the summer solstice
-
-h0 = acos(-tan(phi)*tan(delta));
-q = s0/pi*dist_sun^2*(h0*sin(phi)*sin(delta)+cos(phi)*cos(delta)*sin(h0));
-
-insolation(i) = q;
-end
-
-subplot(6,1,5);
-plot(INSOLN_LA2004_BTL_ASC(1:plot_back,1),insolation(1:plot_back));
-title('insolation')
-
-subplot(6,1,6);
[…]

### 2013-09-15 project001/UVM/paper/setup.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+#
+# set up these files for YOUR disseration!
+#
+# USAGE:
+# ./setup.sh reagan-thesis
+
+OLD_NAMEBASE="storyfinder"
+NEW_NAMEBASE=$1
+
+for FILE in *.tex
+do
+  mv $FILE $(echo $FILE | sed s/$OLD_NAMEBASE/$NEW_NAMEBASE/)
+done
+
+\rm $OLD_NAMEBASE*
+
+sed -i .old s/$OLD_NAMEBASE/$NEW_NAMEBASE/ $NEW_NAMEBASE-revtex4.settings.tex
+
+\rm *.old

### 2013-09-15 project001/UVM/src/paillard_discrete.m

@@ -1,0 +1,47 @@
+function [new_state,curr_state_time,tipI3flag] = paillard_discrete(curr_state,insol,curr_state_time,tstep,tipI3flag,params)
+% could write this model as a handle class
+% i won't
+
+% params = {i0,i1,i2,i3,t_g}
+
+% preset is to stay the same
+new_state = curr_state;
+
+switch curr_state
+    case 'i'
+        % can only transistion to g, if insolation low enough
+        if insol < params{1}
+            new_state = 'g';
+            curr_state_time = -tstep;
+        end
+    case 'g'
+        % can only transition to G, if a bunch of conditions are met
+        % only point of confusion: if i3 breached, do we reset time?
+        %   -going to start with no
+        if insol < params{4} % insolation is low enough
+            if curr_state_time > params{5} % enough time has happened
+                if tipI3flag % and haven't tipped before
+                    new_state = 'G';
+                    curr_state_time = -tstep;
+                else % have tipped before, so need to go under i2
+                    if insol < params{3}
+                        new_state = 'G';
+                        curr_state_time = -tstep;
+                        tipI3flag = 0;
+                    end
+                end
+            end
+        else % insolation got too high, reset timer
+            tipI3flag = 1;
+            curr_state_time = -tstep; % this resets the time
+        end
+    case 'G'
+        % simple: only drop down if insolation gets too big
+        if insol > params{2}
+            new_state = 'i';
+            curr_state_time = -tstep;
+        end
+end
+
+% add to the time
+curr_state_time = curr_state_time + tstep;

### 2013-09-15 project001/UVM/src/run_model.m

@@ -1,0 +1,93 @@
+% run_model.m
+%
+% run model
+
+% load the insolation data as insol_data
+load ../data/insolation/INSOLN.LA2004.BTL.mat
+
+% renormalize the insolation data to zero mean and unit variance
+insol_data(:,6) = insol_data(:,6)-mean(insol_data(:,6)); % zero mean
+insol_data(:,6) = insol_data(:,6)/std(insol_data(:,6)); % unit variance
+
+start_time = -900; % kYr
+tstep = 1; % timestep of our insolation data
+
+initial_state = 'G';
+
+% initialize all of the model variables (constructor...)
+curr_state = initial_state;
+curr_state_time = 0;
+tipI3flag = 0;
+params = {-0.75,0,0,1,33}; % {i0,i1,i2,i3,t_g}; % paillard {-0.75,0,0,1,33}
+
+% to save the data (for plotting)
+state_vec = zeros(abs(start_time),1);
+
+count = 0; % keep track of iterations
+for time = start_time:tstep:0
+    count = count + 1; % ++
+    insol = insol_data(-time+1,6); % load data
+    
+    [curr_state,curr_state_time,tipI3flag] = paillard_discrete(curr_state,insol,curr_state_time,tstep,tipI3flag,params);
+
+    % save the states
+    switch curr_state
+        case 'i'
+            state_vec(count) = 1;
+        case 'g'
+            state_vec(count) = 0;
+        case 'G'
+            state_vec(count) = -1;
+    end
+    
+end
+
+tmpfigh = gcf;
+clf;
+figshape(800,800);
+tmpfilename = 'discrete_plot';
+tmpfilenoname = sprintf('%s_noname',tmpfilename);
+
+set(gcf,'DefaultAxesFontname','helvetica');
+set(gcf,'DefaultLineColor','r');
+set(gcf,'DefaultAxesColor','none');
+set(gcf,'DefaultLineMarkerSize',5);
+set(gcf,'DefaultLineMarkerEdgeColor','k');
+set(gcf,'DefaultLineMarkerFaceColor','g');
+set(gcf,'DefaultAxesLineWidth',0.5);
+set(gcf,'PaperPositionMode','auto');
+
+tmpsym = {'o','s','v','o','s','v'};
+tmpcol = {'b','g',[101 45 93]/255,'r','k','m'};
+
[…]

### 2013-09-16 project001/UVM/paper/mcrn501-p1-uvm.body.tex
added to paper, downloaded and plotted lisiecki d18O data for paper

@@
 Understanding the past climate of the earth allows for us to begin to understand what effects we are having upon it, and to predict what the future will bring.
-From ice cores, a rough temperature record of Earth is constructed and we explain attempt to explain that.
+From the analysis of isotope ratios in ice cores and marine sediment cores, we obtain knowledge of past temperature from the ratio of $\delta ^{18} O$, a proxy for combined temperature and global ice volume.
+This time series is constructed over the past 2 Myr, with higher values indicating a colder climate as the ocean is enriched with $\delta ^{18} O$ during the colder conditions.
 The variation in the orbital parameters of our planet affect the amount of solar insolation recieved, and the timescale variation of these quantities is one possible forcing for past climatic change.
-The eliptical elements of the orbit are changing slowly in time, as is the orientation of the planet's spin axis.
-A three state model from Paillard fits the data emprically, and we attempt a rigorous fit of this model to data using the tools of evolutionary computation.
+This was hypothesized as early as the nineteenth century, and further elucidated by Milankovitch, whose name the study of these variations bears.
+The eccentricity of our eliptical orbit is changing slowly in time, as is the orientation of the planet's spin axis and the longitude closest to the sun at the vernal equinox.
+We refer to these parameters as $e$, $\epsilon$ and $\overline{\omega}$, respectively.
+
+The main questions arising about this time period from Dijkstra \cite{dijkstra2013} that any satisfactory theory of the Pleistocene Ice Ages should explain are:
+\begin{itemize}
+\item Why did the glacial-interglacial cycles appear in the Pleistocene?
+\item Which processes in the climate system caused the glacial-interglacial changes in global mean temperature, $p^a_{CO_2}$, and ice sheet extent?
+\item What caused the transistion (the MPT) from the 41-kyr world to the 100-kyr world about 700 kyr ago?
+\end{itemize}
+
+\begin{figure}[tpb!]
+\centering
+  \includegraphics[width=.48\textwidth]{../data/dO18/d18O_lisiecki_noname.pdf}
+  \caption{
+    The LR04 benthic $\delta ^{18} O$ stack over the Pleistocene, constructed by the graphic correlation of fifty-seven globally distributed benthic $\delta ^{18} O$ records versus time (data from Lisieki \cite{lisiecki2005pliocene}).
+  }
+  \label{fig:insol-data}
+\end{figure}
+
+The Milankovitch theory of orbital forcing is now the `null hypothesis' answer to these questions, and various models are available that relate this orbital forcing to global mean temperature and ice sheet extent have been proposed.
+The fit between the changing solar insolation and $\delta ^{18} O$ has been demonstrated not to be linear, opening the door for more complex explanations.
+The potentially impoortant feedbacks of ice-albedo, height-mass balance, load-accumulation, and precipation-temperature have all been combined into quasi-equilibrium and internal oscillation models, with the latter successfully demonstrating a reasonable fit to the data.
+In addition to these more complex models is a switching equilibria model from Didier Paillard that fits the data emprically, and does ``remarkable'' considering the simplicity of the model.
+
+In the present work we attempt a rigorous fit of Paillard's model to data using the tools of evolutionary computation.
+
+\begin{figure}[tpb!]
+\centering
+  \includegraphics[width=.45\textwidth]{../src/discrete_plot_noname.pdf}
+  \caption{
+    The discrete Paillard model.
+  }
+  \label{fig:paillard-fig3}
+\end{figure}
 
@@
 
+\begin{figure}[tpb!]
+\centering
+  \includegraphics[width=.48\textwidth]{../data/insolation/laskar2004/insol_data2_noname.pdf}
+  \caption{
+    The average daily solar insolation $\overline{Q}_{\text{day}}$ is computed on the summer solstice at 65N with the prodived $e,\epsilon,$ and $\overline{\omega}$.
+  }
+  \label{fig:insol-data}
+\end{figure}
+
 \section{Evolutionary Algorithm Methods}
@@
 Figure \ref{fig:insol-data} depicts the reconstruction of insolation data back in time.
-
-\begin{figure}[tpb!]
-\centering
-  \includegraphics[width=.48\textwidth]{../data/insolation/insol_data2_noname.pdf}
-  \caption{
-    The average daily solar insolation $\overline{Q}_{\text{day}}$ is computed on the summer solstice at 65N with the prodived $e,\epsilon,$ and $\overline{\omega}$.
-  }
-  \label{fig:insol-data}
-\end{figure}
 
@@
 
-\begin{figure}[tpb!]
-\centering
-  \includegraphics[width=.45\textwidth]{../src/discrete_plot_noname.pdf}
-  \caption{
-    The discrete Paillard model.
-  }
-  \label{fig:paillard-fig3}
-\end{figure}
-
 \section{Conclusion}

### 2013-09-16 project001/UVM/data/dO18/plot_data.m

@@ -1,0 +1,78 @@
+close all;
+clear all;
+%% make pretty plot
+
+tmpfigh = gcf;
+clf;
+figshape(600,600);
+tmpfilename = 'd18O_lisiecki';
+tmpfilenoname = sprintf('%s_noname',tmpfilename);
+
+%set(gcf,'Color','none');
+
+set(gcf,'DefaultAxesFontname','helvetica');
+set(gcf,'DefaultLineColor','r');
+set(gcf,'DefaultAxesColor','none');
+set(gcf,'DefaultLineMarkerSize',5);
+set(gcf,'DefaultLineMarkerEdgeColor','k');
+set(gcf,'DefaultLineMarkerFaceColor','g');
+set(gcf,'DefaultAxesLineWidth',0.5);
+set(gcf,'PaperPositionMode','auto');
+
+tmpsym = {'o','s','v','o','s','v'};
+tmpcol = {'b','g',[101 45 93]/255,'r','k','m'};
+
+positions(1).box = [.2 .2 .7 .7];
+axesnum = 1;
+tmpaxes(axesnum) = axes('position',positions(axesnum).box);
+
+plot_back=1251;
+
+tmp = load('LR04stack.csv');
+
+plot(fliplr(-tmp(1:plot_back,1)),fliplr(-tmp(1:plot_back,2)),'Color',tmpcol{1},'LineWidth',3);
+
+
+
+tmpxlab=xlabel('ky from J2000','fontsize',25,'verticalalignment','top','fontname','helvetica','interpreter','latex');
+tmpylab=ylabel('$\delta ^{18}$ O','fontsize',25,'verticalalignment','bottom','fontname','helvetica','interpreter','latex');
+
+% disp(get(tmpxlab,'position'))
+% 
+set(tmpxlab,'position',get(tmpxlab,'position') - [0 .05 0]); % [x y 0]
+set(tmpylab,'position',get(tmpylab,'position') - [.05 0 0]);
+
+%set(gca,'xtick',months_cum_rank)
+%month_names={'J','F','M','A','M','J','J','A','S','O','N','D'};
+%set(gca,'xticklabel',month_names)
+
+%ylim([5.5 3])
+set(gca,'ytick',fliplr([-3 -3.5 -4 -4.5 -5 -5.5]));
+tmptmp = {'' '5'  '' '4' '' '3'};
+set(gca,'yticklabel',tmptmp,'fontsize',16,'fontname','helvetica');
+
[…]

### 2013-09-17 project001/UVM/paper/methods.tex
adding Morgan's stuff

\documentclass[onecolumn]{article}
\hoffset=0in
\voffset=-.2in
\marginparwidth=0in
\marginparsep=0in
\oddsidemargin=0in
\topmargin=0in
\headheight=0in
\textwidth=470pt
\textheight = 620pt
\fontsize{12}{15}
\pagestyle{empty}
\usepackage{multicol}
\usepackage{cite}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{amsmath,amssymb}
\usepackage{mathptm,graphicx,rotate,color}
\newcommand{\bb}[1]{\mathbb{#1}}
\usepackage{cite}
\newcommand{\DO}{$\delta^{18}O$ }
\begin{document}
\section{Methods}
\indent We attempt to search the parameter space within the limits suggested by Paillard \cite{paillard} in hopes of better fitting the both the discrete and the ODEE models to the \DO data by using a genetic algorithm (GA) and covariance matrix adaption evolutionary scheme (CMA-ES) to evolve possible parameters. CMA-ES and GA are well-known and documented evolutionary algorithms for evolving real-valued vectors that minimize a particular fitness function when evaluated. For this investigation, we chose to minimize the root mean square error of the normalized model compared to the \DO data. \\
\subsection{Datasets}
\indent We obtained to necessary data sets for running our experiments. Our \DO data comes from Tiedermann et al. \cite{d18} as the result of analyzing an ice core sample taken from up to 273.80 m in 1986. A simple linear transformation is used to scale from depth to the age of the samples. The ice core sample was taken in the North Atlantic Ocean. While this may not be the exact dataset used by Paillard, we believe the data contained here should be sufficiently similar and relevant in our efforts to fit Paillard's models to historical climate behavior.\\
\indent We employ the same insolation dataset used by Paillard in \cite{paillard}; namely, we use the data provided by A. Berger in \cite{berger}. Berger obtains this curve representing the earth's atmospheric insulation about the latitude $65^{o}N$ by modeling the earth's rotation about the sun. His calculations account for such variables as the angle of the earth's axis from perpendicular to the sun, and eccentricity. This insolation curve acts as the forcing for our model.\\
\subsection{Models}
\indent Paillard \cite{paillard} suggests two models for obtaining the \DO curves from atmospheric insolation data. These models represent the transformations of three different glacial states: $i$ (interglacial), $g$ (mild glacial), $G$ (full glacial). Both models allow us to impose limits on the insolation, the ice volume that determines the glacial state, and the expected time spent in each glacial state. In both models, it is assumed that the only three possible transitions exists. A $i\rightarrow g$ (interglacial$\rightarrow$ mild glacial) transition occurs when when the insolation falls below a parameter $i_{0}$. A $g\rightarrow G$ (mild glacial $\rightarrow$ full glacial) transition occurs when a maximum ice volume $v_{max}$ is exceeded. Finally, a $G\rightarrow i$ (full glacial$\rightarrow$ interglacial) transition occurs when the insolation grows above a parameter. \\
\indent Paillard first suggests a discrete model for obtaining the \DO curve from insolation data. This model has four insolation parameters.  $i_{0}$ and $i_{1}$ represent the insolation values required for $i-g$ and $G-i$ transitions to occur, respectively. The other insolation parameters $i_{2}$ and $i_{3}$ identify a range of possible forcing values that would allow a $g-G$ transition to occur. A further parameter, $\tau_{g}$, is a time constant determining the minimum number of years for the model to spend in a mild glacial state. In an effort to reduce the search space, we adopt Paillard's assumption of $\tau_{g}=33$ kyr for our discrete model and attempt to evolve values for the parameters $i_{0}$, $i_{1}$, $i_{2}$, and $i_{3}$.\\
\indent An ODE model is also presented by Paillard, which is given by \[ \frac{dv}{dt}=\frac{v_{R}-v}{\tau_{R}}-\frac{F}{\tau_{F}} \] where $v$ is the ice volume, $v_{R}$ and $\tau_{R}$ are the volume and the time constants, respectively, corresponding to the glacial state. $\tau_{F}$ is a time constant associated with a forcing parameter $F$. The forcing parameter $F$ is given through a truncation of the insolation data described by \[ F=\frac{1}{2}(x+\sqrt{4a^{2}+x^{2}})\] with $x$ being the insolation and a tunable compression parameter $a$. $i_{0}$ and $i_{1}$ are the same as in the discrete model, and the other insolation parameters are omitted. Again, we reduce our search space by adopting the assumption presented by Paillard that $\tau_{i}=10$ kyr, $\tau_{g}=\tau_{G}=50$ kyr, $\tau_{F}=25$ kyr, and $a=1$. We attempt to provide good choices for the remaining insolation parameters $i_{0}$ and $i_{1}$.\\
\subsection{Evolutionary Algorithms}
\indent Both of the evolutionary algorithms we used are designed to search through the space $\bb{R}^{n}$ with a population of $n$ length real-valued vectors with each entry corresponding to a parameter in our model. Through common genetic mechanisms, like reproduction and mutation, we are able to evolve vectors representing potentially good parameter choices for the insolation parameters in each model to fit the \DO  data. Both algorithms require a fitness function with which to assess how well the parameter choices represented by an individual vector represent the \DO data through the models. We employ a fitness function which integrates the model and then attempts to minimize the root mean square error of the resulting time series compared to the true \DO data. This fitness function allows us to apply selection pressure on the populations of vectors that results in competition within the population to fit the \DO data the best.\\
\vspace{.25cm}\\
\noindent{\bf Covariance Matrix Adaption Evolutionary Scheme (CMA-ES)}\\
\indent CMA-ES is often a good initial choice of evolutionary algorithm for evolving real-valued vectors because the algorithm itself is not expensive and the algorithm often yields a reasonably good result after only a few generations. The algorithm begins with a uniformly distributed cloud of points in $\bb{R}^{n}$. Each point represents parameter choices for the models. We asses the fitness of each point and create a covariance matrix weighted by the fitness values. By taking the average of the fitness values, we can determine weighted center of the cloud of points, and we can determine a direction to move that point that will hopefully yield a better average fitness. Having relocated the average point, we create a new cloud of points around the average, however we now use the covariance matrix to determine how to distribute the individuals points. For example, if the population members all agree at a particular index, then they will have small covariances. This will caus!
 e the new cloud of points to be very narrow in that dimension since all of the population members are in agreement. On the other hand, if all of the individuals disagree in a particular dimension, then there will be large covariances representing a need to explore more in that dimension. Thus, the cloud of points should be wider in that dimension. We fix the number of generations to be 500 for our experiments. More information about CMA-ES can be found in \cite{suttorp}.\\
\begin{center}
	\includegraphics[scale=.6]{cmaes.png}
\end{center}
Figure 1. A cartoon representing 6 generations in a CMA-ES experiment. Here, lighter colors indicate improved fitness. The dashed lines represent the influence of the covariance matrix.\\
\vspace{.25cm}\\
\noindent{\bf Genetic Algorithm (GA)}\\
\indent GA's are typically more expensive than CMA-ES, but allow for an increased wealth in ways for population members to combine information. GA's often yield greater runtimes than CMA-ES but can also yield more robust explorations of parameter space. The algorithm begins with a population of real-valued vectors. Each vector is assessed for fitness. The fitter individuals in the population are chosen as candidates for reproduction in the form of crossover. Here, we use uniform crossover, which means that given two ``parent" vectors from the population, we create a ``child" vector by randomly selecting indices to contain the value stored in that index by parent 1. The remaining indices are filled with the values stored in parent 2 at those indices. A second child vector is created by taking the same random indices, and swapping the roles of the parents. These two children vectors then replace their parents in the population. Individuals in the population are also subject to!
  mutation. Mutation is performed on individuals where a small amount is added or subtracted from each index in an individual vector with some small probability. Crossover often leads the population to converge on good solutions, while mutation generally causes the population to explore parameter space. Striking a balance between these two forces is vital and often very difficult depending on the noise in the fitness landscape; that is, the fitness values associated with each point in the parameter space. For our experiments, we use a population size of 500 and calculate 500 generations.\\
\begin{center}
	\includegraphics[scale=.6]{gaImg.png}
\end{center}
Figure 2. A schematic explaining the basic steps in a genetic algorithm.\\
\section{Results}
\indent Having run 100 CMA-ES and GA experiments, we discuss the fit of the best solutions here. Firstly, CMA-ES underperformed. None of the CMA-ES experiments yielded results worth noting here. This is likely due to unresolvable noise in the fitness landscape. We focus instead on the more promising results from our GA experiments.\\
\begin{center}
$\begin{array}{cc}
	\includegraphics[width=.5\columnwidth]{discrete}&\includegraphics[width=.5\columnwidth]{ode}
\end{array}$
\end{center}
Figure 3. {\bf (Left)} The best result from the GA experiments attempting to fit parameters to Paillard's discrete model. {\bf (Right)} THe best result from the GA experiments attempting to fit parameters to Paillard's ODE model. The root mean square errors are reported at the top. \\
\vspace{.25cm}\\
\indent Figure 3 demonstrates the results of our most promising GA experiments along with their root mean square errors (RMSE) from the \DO data. We should note that the particular parameter choices proposed by Paillard in \cite{paillard} yield RMSE$=1.42$ for the discrete model, and RMSE$=1.7186$ for the ODE model. Figure 3 shows that our GA experiment ($i_{0}= -0.7293$, $i_{1}=-0.2287$, $i_{2}=0.0803$, and $i_{3}=0.9862$) was only narrowly outperformed by Paillard's parameter choices for the discrete model, and that our GA experiment ($i_{0}= -0.9677$ and $i_{1}=0.0294$) yields a better fitting ODE model than the model suggested by Paillard's parameter choices for this model.\\
\indent 
\begin{thebibliography}{9}
	\bibitem{paillard}
	Paillard D. Laboratoire de Modélisation du Climat et de l'Environnement, CEA/DSM, Centre d'Etudes de Saclay, 91191 Gif-sur-Yvette, France
	\bibitem{d18}  Tiedemann, R et al. (1994): Stable oxygen isotope ratios of Cibicidoides wuellerstorfi from ODP Hole 108-659A in the North Atlantic. doi:10.1594/PANGAEA.52517,
        In Supplement to: Tiedemann, Ralf; Sarnthein, Michael; Shackleton, Nicholas J (1994): Astronomic timescale for the Pliocene Atlantic d18O and dust flux records of Ocean Drilling Program site 659. Paleoceanography, 9(4), 619-638, doi:10.1029/94PA00208
        \bibitem{berger} Berger, A. Long-term variations of daily insolation and Quaternary climatic change.J.Atmos. Sci. 35,
2362–2367 (1978).
	\bibitem{suttorp}
T. Suttorp, N. Hansen, C. Igel, Efficient covariance matrix update for
variable metric evolution strategies, Machine Learning 75 (2009) 167–
197
\end{thebibliography}
\end{document}

### 2013-09-17 project001/UVM/paper/mcrn501-p1-uvm.body.tex
presentation and paper finished for today's class

@@
 
-Morgan
+\newcommand{\bb}[1]{\mathbb{#1}}
+\newcommand{\DO}{$\delta^{18}O$ }
+
+We attempt to search the parameter space within the limits suggested by Paillard \cite{paillard1998timing} in hopes of better fitting the both the discrete and the ODEE models to the \DO data by using a genetic algorithm (GA) and covariance matrix adaption evolutionary scheme (CMA-ES) to evolve possible parameters.
+CMA-ES and GA are well-known and documented evolutionary algorithms for evolving real-valued vectors that minimize a particular fitness function when evaluated.
+For this investigation, we chose to minimize the root mean square error of the normalized model compared to the \DO data.
+
+\subsection{Datasets for fitting}
+We obtained to necessary data sets for running our experiments.
+Our \DO data comes from Tiedermann et al.
+\cite{d18} as the result of analyzing San ice core sample taken from up to 273.80 m in 1986.
+A simple linear transformation is used to scale from depth to the age of the samples.
+The ice core sample was taken in the North Atlantic Ocean.
+While this may not be the exact dataset used by Paillard, we believe the data contained here should be sufficiently similar and relevant in our efforts to fit Paillard's models to historical climate behavior.\\
+\indent We employ the same insolation dataset used by Paillard in \cite{paillard1998timing}; namely, we use the data provided by A. Berger in \cite{berger1981long}.
+Berger obtains this curve representing the earth's atmospheric insulation about the latitude $65^{o}N$ by modeling the earth's rotation about the sun.
+His calculations account for such variables as the angle of the earth's axis from perpendicular to the sun, and eccentricity.
+This insolation curve acts as the forcing for our model.
+
+\subsection{Models in more detail}
+Paillard \cite{paillard1998timing} suggests two models for obtaining the \DO curves from atmospheric insolation data.
+These models represent the transformations of three different glacial states: $i$ (interglacial), $g$ (mild glacial), $G$ (full glacial).
+Both models allow us to impose limits on the insolation, the ice volume that determines the glacial state, and the expected time spent in each glacial state.
+In both models, it is assumed that the only three possible transitions exists.
+A $i\rightarrow g$ (interglacial$\rightarrow$ mild glacial) transition occurs when when the insolation falls below a parameter $i_{0}$.
+A $g\rightarrow G$ (mild glacial $\rightarrow$ full glacial) transition occurs when a maximum ice volume $v_{max}$ is exceeded.
+Finally, a $G\rightarrow i$ (full glacial$\rightarrow$ interglacial) transition occurs when the insolation grows above a parameter.
+
+Paillard first suggests a discrete model for obtaining the \DO curve from insolation data.
+This model has four insolation parameters.
+$i_{0}$ and $i_{1}$ represent the insolation values required for $i-g$ and $G-i$ transitions to occur, respectively.
+The other insolation parameters $i_{2}$ and $i_{3}$ identify a range of possible forcing values that would allow a $g-G$ transition to occur.
+A further parameter, $\tau_{g}$, is a time constant determining the minimum number of years for the model to spend in a mild glacial state.
+In an effort to reduce the search space, we adopt Paillard's assumption of $\tau_{g}=33$ kyr for our discrete model and attempt to evolve values for the parameters $i_{0}$, $i_{1}$, $i_{2}$, and $i_{3}$.
+
+An ODE model is also presented by Paillard, which is given by \[ \frac{dv}{dt}=\frac{v_{R}-v}{\tau_{R}}-\frac{F}{\tau_{F}} \] where $v$ is the ice volume, $v_{R}$ and $\tau_{R}$ are the volume and the time constants, respectively, corresponding to the glacial state.
+$\tau_{F}$ is a time constant associated with a forcing parameter $F$.
+The forcing parameter $F$ is given through a truncation of the insolation data described by \[ F=\frac{1}{2}(x+\sqrt{4a^{2}+x^{2}})\] with $x$ being the insolation and a tunable compression parameter $a$.
+$i_{0}$ and $i_{1}$ are the same as in the discrete model, and the other insolation parameters are omitted.
+Again, we reduce our search space by adopting the assumption presented by Paillard that $\tau_{i}=10$ kyr, $\tau_{g}=\tau_{G}=50$ kyr, $\tau_{F}=25$ kyr, and $a=1$.
+We attempt to provide good choices for the remaining insolation parameters $i_{0}$ and $i_{1}$.\\
+
+\subsection{Evolutionary Algorithms}
+Both of the evolutionary algorithms we used are designed to search through the space $\bb{R}^{n}$ with a population of $n$ length real-valued vectors with each entry corresponding to a parameter in our model.
+Through common genetic mechanisms, like reproduction and mutation, we are able to evolve vectors representing potentially good parameter choices for the insolation parameters in each model to fit the \DO  data.
+Both algorithms require a fitness function with which to assess how well the parameter choices represented by an individual vector represent the \DO data through the models.
+We employ a fitness function which integrates the model and then attempts to minimize the root mean square error of the resulting time series compared to the true \DO data.
+This fitness function allows us to apply selection pressure on the populations of vectors that results in competition within the population to fit the \DO data the best.\\
+\vspace{.25cm}
+
+\noindent{\bf Covariance Matrix Adaption Evolutionary Scheme (CMA-ES)}\\
+CMA-ES is often a good initial choice of evolutionary algorithm for evolving real-valued vectors because the algorithm itself is not expensive and the algorithm often yields a reasonably good result after only a few generations.
+The algorithm begins with a uniformly distributed cloud of points in $\bb{R}^{n}$.
+Each point represents parameter choices for the models.
+We asses the fitness of each point and create a covariance matrix weighted by the fitness values.
+By taking the average of the fitness values, we can determine weighted center of the cloud of points, and we can determine a direction to move that point that will hopefully yield a better average fitness.
+Having relocated the average point, we create a new cloud of points around the average, however we now use the covariance matrix to determine how to distribute the individuals points.
+For example, if the population members all agree at a particular index, then they will have small covariances.
+This will cause the new cloud of points to be very narrow in that dimension since all of the population members are in agreement.
+On the other hand, if all of the individuals disagree in a particular dimension, then there will be large covariances representing a need to explore more in that dimension.
+Thus, the cloud of points should be wider in that dimension.
+We fix the number of generations to be 500 for our experiments.
+More information about CMA-ES can be found in \cite{suttorp}.
+
+\begin{figure}[tpb!]
+\centering
+  \includegraphics[width=.48\textwidth]{cmaes.png}
+  \caption{
+    A cartoon representing 6 generations in a CMA-ES experiment. Here, lighter colors indicate improved fitness. The dashed lines represent the influence of the covariance matrix.\\
+    }
+  \label{fig:cmaes}
+\end{figure}
+
+\noindent{\bf Genetic Algorithm (GA)}\\
+GA's are typically more expensive than CMA-ES, but allow for an increased wealth in ways for population members to combine information.
+GA's often yield greater runtimes than CMA-ES but can also yield more robust explorations of parameter space.
+The algorithm begins with a population of real-valued vectors.
+Each vector is assessed for fitness.
+The fitter individuals in the population are chosen as candidates for reproduction in the form of crossover.
+Here, we use uniform crossover, which means that given two ``parent" vectors from the population, we create a ``child" vector by randomly selecting indices to contain the value stored in that index by parent 1.
+The remaining indices are filled with the values stored in parent 2 at those indices.
+A second child vector is created by taking the same random indices, and swapping the roles of the parents.
+These two children vectors then replace their parents in the population.
+Individuals in the population are also subject to mutation.
+Mutation is performed on individuals where a small amount is added or subtracted from each index in an individual vector with some small probability.
+Crossover often leads the population to converge on good solutions, while mutation generally causes the population to explore parameter space.
+Striking a balance between these two forces is vital and often very difficult depending on the noise in the fitness landscape; that is, the fitness values associated with each point in the parameter space.
+For our experiments, we use a population size of 500 and calculate 500 generations.
+
+\begin{figure}[tpb!]
+\centering
+  \includegraphics[width=.48\textwidth]{gaImg.png}
+  \caption{
+      A schematic explaining the basic steps in a genetic algorithm.
+    }
+  \label{fig:ga}
+\end{figure}
 
@@
 
-Figure \ref{fig:insol-data} depicts the reconstruction of insolation data back in time.
+Constructin the insolation from orbital parameters, Figure \ref{fig:insol-data} depicts the reconstruction of insolation data back in time from Laskar's data set.
+And verifying that our interpretation of the model is the same as Paillard intended, Figure \ref{fig:paillard-fig3} is a direct representation of the second figure of Paillard's original paper \cite{paillard1998timing}.
 
-Figure \ref{fig:paillard-fig3} is a direct representation of the second figure of Paillard's original paper \cite{paillard1998timing}.
+Having run 100 CMA-ES and GA experiments, we discuss the fit of the best solutions here.
+Firstly, CMA-ES underperformed.
+None of the CMA-ES experiments yielded results worth noting here.
+This is likely due to unresolvable noise in the fitness landscape.
+We focus instead on the more promising results from our GA experiments.
 
@@
 \centering
-  \includegraphics[width=.53\textwidth]{paillardFig3.pdf}
+  \includegraphics[width=.48\textwidth]{../src/results_plot_noname.pdf}
   \caption{
-    A first run of the Paillard ODE model, over the $\delta ^{18} O$ record show a reasonable emprical fit.
+      The best result from the GA experiments attempting to fit parameters to Paillard's discrete and ODE model, with the \DO data used for fitness.
     }
-  \label{fig:paillard-fig3}
+  \label{fig:results}
 \end{figure}
+
+Figure \ref{fig:results} demonstrates the results of our most promising GA experiments along with their root mean square errors (RMSE) from the \DO data.
+We should note that the particular parameter choices proposed by Paillard in \cite{paillard1998timing} yield RMSE$=1.42$ for the discrete model, and RMSE$=1.7186$ for the ODE model.
+Figure 3 shows that our GA experiment ($i_{0}= -0.7293$, $i_{1}=-0.2287$, $i_{2}=0.0803$, and $i_{3}=0.9862$) was only narrowly outperformed by Paillard's parameter choices for the discrete model, and that our GA experiment ($i_{0}= -0.9677$ and $i_{1}=0.0294$) yields a better fitting ODE model than the model suggested by Paillard's parameter choices for this model.
 
@@
 
-The fitted Paillard-Reagan-Frank model of climatic variability wholly explains the temperature variation in the $\delta ^{18} O$ record.
+The fitted Paillard model of climatic variability makes a strong case for differing atmospheric states playing a role in the temperature variation in the $\delta ^{18} O$ record.
+The methods of evolutionary computation are important in climate science to assist in model validation, an important step in understanding.
 With only forcing from changing solar insolation, this verifies that past climatic varaition of larger time scales is a result of the variation of the Earth's orbital parameters.

### 2013-09-17 project001/UVM/src/results_plot.m

@@ -1,0 +1,87 @@
+% plot results
+
+
+load ../data/dO18/d18o.txt
+load odeData.txt
+load discreteData.txt
+
+tmpfigh = gcf;
+clf;
+figshape(800,800);
+tmpfilename = 'results_plot';
+tmpfilenoname = sprintf('%s_noname',tmpfilename);
+
+set(gcf,'DefaultAxesFontname','helvetica');
+set(gcf,'DefaultLineColor','r');
+set(gcf,'DefaultAxesColor','none');
+set(gcf,'DefaultLineMarkerSize',5);
+set(gcf,'DefaultLineMarkerEdgeColor','k');
+set(gcf,'DefaultLineMarkerFaceColor','g');
+set(gcf,'DefaultAxesLineWidth',0.5);
+set(gcf,'PaperPositionMode','auto');
+
+tmpsym = {'o','s','v','o','s','v'};
+tmpcol = {'b','g',[125, 38, 205]/255,'r','k','m'};
+
+positions(1).box = [.2 .2 .7 .7];
+axesnum = 1;
+tmpaxes(axesnum) = axes('position',positions(axesnum).box);
+
+plot(odeData(:,1),(odeData(:,2)-mean(odeData(:,2)))./max(odeData(:,2)-mean(odeData(:,2)))+1,'Color',tmpcol{1},'LineWidth',3)
+hold on;
+plot(discreteData(:,1),(discreteData(:,2)-0.2785)./(max(discreteData(:,2)-0.2785)+.3)+3,'Color',tmpcol{3},'LineWidth',3)
+plot(d18o(:,1),(d18o(:,2)-mean(d18o(:,2)))./max(d18o(:,2)-mean(d18o(:,2)))+5,'Color',tmpcol{2},'LineWidth',3)
+
+ylim([-.2 6])
+xlim([-1000 20])
+% shift = 3.5;
+set(gca,'ytick',[0 1 2 4 5 6])
+set(gca,'yticklabel',{-1,0,1,-1,0,1})
+
+% line(1) = plot(fliplr(insol_data(1:-start_time,1)),params{4}.*ones(length(1:-start_time),1)+shift,'b--','LineWidth',1.1);
+% line(2) = plot(fliplr(insol_data(1:-start_time,1)),params{1}.*ones(length(1:-start_time),1)+shift,'r--','LineWidth',1.2);
[…]

### 2013-09-19 project001/UVM/data/insolation/laskar2004/plot_data.m
updated for class on 09-19

@@ -107,6 +107,9 @@ data = {tmp(1:plot_back,2),tmp(1:plot_back,3),sin(tmp(1:plot_back,4))...
     ,tmp(1:plot_back,2).*sin(tmp(1:plot_back,4)),insolation(1:plot_back)};
 insol_data = [tmp(:,1) tmp(:,2) tmp(:,3) sin(tmp(:,4))...
      tmp(:,2).*sin(tmp(:,4)) insolation(:)];
+% without the sin
+insol_data2 = [tmp(:,2) tmp(:,3) tmp(:,4)...
+     insolation(:)];
  
 ranges = zeros(2,5);
 

### 2013-11-03 project002/ipcc-andy.tex
updating from class

\subsection{Sources of uncertainty}

In responce to increasing SST, the changes in climate tend to agree in a handful of observed ``convergence zones.''
Convergence zones are long and narrow bands of tropical convection anchored by SST structures.
These changes include a wet-get-wetter paradigm and a warmer-get-wetter mechanism, for uniform and projected SST increases, respectively.
The spatial pattern of SST change is the main driver of uncertainty between these two modes of precipitation change in the observed convergence zones.
SST change and preciptiation plotted over one another can be seen in Figure \ref{fig:148}.\\

\begin{figure*}[h!]
\begin{center}
  \includegraphics[width=0.790\textwidth]{fig148.png}
  \label{fig:148}
\end{center}
\end{figure*}

Over the tropical Atlantic, models agree poorly over current conditions due mainly to a northernly biased ITCZ and a cold tongue in (our) summer.
Models are able to capture the general warming trend over the past century in the Atlantic, but this results in a southward displaced Atlantic ITCZ.\\

Observational uncertainty, limited data, and changing measurement techniques all make the projection of zonal SST gradient difficult, and some models show strengthening while other models show weakening.
The ENSO variability over the past century is reproduced both in variability and spatial pattern by models with external forcing and without forcing, such that there is little agreement on whether external forming plays a role in the observed changes in ENSO.\\

Again due to uncertainties in SST projections, the ENSO spatial pattern change has low confidence.\\

\subsection{Agreement in convergence zones}

The three main convergence zones are the ITCZ, the South Pacific Convergence Zone (SPCZ), and the South Atlantic Covergence Zone (SACZ).\\
%%Of these, the IPCZ has known biases which are strongest in (our) summer, while the SPCZ and the SACZ model predictions generally agree over the past century.

The ITCZ suffers an unrealistic double-ITCZ pattern over the tropical Pacific and the Atlantic, which results in excessive rainfall over the southern flank.
This is due to the incorrect orientation of the SPCZ, missing the southeastward orientation of the SPCZ and aligning it with the ITCZ.
Changes in ENSO displace and change the orientation of the SPCZ as well, moving the SPCZ northeast and aligning it zonally in response to stronger events.
This shift to zonally-oriented SPCZ is consistent in both the CMIP3 and CMIP5 models, and increasing in frequency in the future.
This relies on a reduction in near-equatorial SST gradient, a consitent finding in response to human forcing.
A consistent shift of the SPCZ woudl result in major changes for the southwest Pacific, including potentially a longer dry spell.\\

In the SACZ, changes have a major impact on the precipitation over southeast South America.
The projected southward displacement and intensification will likely result in changes in the flood and dry conditions of South America, to which the current SPCZ pattern effects over Brazil.\\

\subsection{ENSO persistence}

Depsite the uncertainties in the spaital particularities of the ENSO, the amplitude modulation are well reproduced over the past century by models and the generation of the 10-100 year modulations by coupled GCM's without forcing.
Poor observations of the detailed coupled air-ocean feedbacks before 1970 make the discernment of forcings' role in ENSO uncertain, as mentioned earlier.\\

The improvement in the modelling of ENSO intensity from CMIP3 to CMIP5 is significant, and due to increased moisture with rainfall variability is expected to increase regionally.
Although the future changes in intensity are not confident due to the many factors on which ENSO depends, we remain very likely that this will continue to be the dominant mode of climate variability.
In Figure \ref{fig:1414} we can see how likely this is.\\

\begin{figure*}[h!]
\begin{center}
  \includegraphics[width=0.490\textwidth]{figure1414.png}
  \caption{Standard deviation of Niño3 SST anomalies from CMIP5 model experiments.
    PI is the pre-industrial control experiment, 20C is the 20th Century control experiment, RCP4.5/8.5 are the 21st Century projections.
    The cross mark on 20C is the current observation.}
  \label{fig:1414}
\end{center}
\end{figure*}

\subsection{Future goals for Climate Scientists}

To make better predictions of the climate, we need to focus on our ability to reproduce the known large scale structures such as covergence zones that drive regional climate changes.
A better understanding of these structures, coupled with better observations, will allow us to ensure that the models accurately represent these structures togay and how they have changed during our observations, such that we can have confidence in future predictions.
In addition, the convergent structures drive regional precipitation patterns, which are economically and socially important outputs of climate modelling for the future. \\

A case study for this work has been the work of the past few decades on the ENSO.
Once poorly understood, it was foreseen by Charney and the vast amounts of work we have put forth to understand this phenomena have allowed us to be able to predict how it will change in the future.
Only with better observations, over a longer time, can we be more sure of the changes in intensity of these structures as well.\\

### 2013-11-03 project002/ipcc-isaac.tex
updating from class

\subsection{Modelling Challenges}

There are fundamental challenges that currently limit the effectiveness of modelling tropical phenomena.  The current generation of climate models (CMIP5) do not consistently predict the same results for particular phenomena.  These phenomena include precipitation patterns, SST patterns, Madden-Julian Oscillation and ENSO.  Whenever models disagree on a specific aspect of any of these phenomena, it makes future prediction must less certain.  Therefore, it is necessary to identify the deficiencies in modelling which cause the discrepancies in model results.

\begin{figure*}[h!]
\begin{center}
  \includegraphics[width=0.790\textwidth]{isaac-fig1.png}
  \label{fig:isaac1}
\end{center}
\end{figure*}

Rainfall projections are an important aspect of tropical modelling.  According to the IPCC report, there is a wide range of inter-model variability in future rainfall projections.  Obvious model differences can be seen in Figure \ref{fig:isaac1} above.  A third of the variability was shown to be caused by differences in the SST warming pattern (Ma and Xie, 2013).  According to Ma and Xie, there is a strong correlation between warmer-than-average SST and increased precipitation within regions.  Thus modeling SST warming patterns is very important for predicting future rainfall.  While SST is projected to increase in general, the pattern of warming is not projected to be uniform.  Further, this pattern of warming is not consistent among models.  Due to the correlation of SST warming and precipitation, any inaccuracy in predicting SST warming patterns is inherited by the modelling of future rainfall.  This is a modelling challenge because SST pattern formation involves complicated processes such as ocean-atmosphere interactions and the interaction between convection and large-scale environment which are not well-understood.\\

\begin{figure*}[h!]
\begin{center}
  \includegraphics[width=0.790\textwidth]{isaac-fig2.png}
  \label{fig:isaac2}
\end{center}
\end{figure*}

The study of the the Intertropical Convergence Zone (ITCZ) is another modeling pursuit that is very relevant to the projection of future precipitation in the tropics.  A major issue in modelling this feature is the appearance of an unrealistic ‘double ITCZ’ pattern in many models.  \\
An example of this ‘double ITCZ’ is seen in the eastern Pacific ocean, where models project a higher level of precipitation both North and South of the equator, as opposed to reality, where a diagonally oriented convergence zone is present in the south.  This can be seen in Figure \ref{fig:isaac2} above. In a study, Brown used a regime sorting technique, in which precipitation change is decomposed into dynamic, thermodynamic, and covariant components (Brown et al., 2012).  This study showed that, in the regions where models most often disagree with reality, the dynamic and thermodynamic components of the precipitation change were also most different from reality.  This suggested that reduced atmospheric moisture and reduced ascent are to explain for the apparent bias.  In a previous study by Brown, it was shown that models that use flux adjustments are the most accurate in reproducing the observed spatial ITCZ pattern, while models that do not use flux adjustments fail to capture the diagonal orientation (Brown et al., 2011).\\

This result suggest the absence of ocean-heat flux adjustments in most CMIP5 models may be responsible for their inaccuracy in modeling the observed ITCZ pattern.  Once again, the influence of SST warming, and in this case zonal SST gradient, is an important factor in modeling another climate phenomena.\\

Considering the above result, one might wonder why more models do not use ocean-atmosphere flux adjustments, since doing so enables better modelling of the ITCZ.  The answer is there is a tradeoff.  The inclusion of ocean-atmosphere flux adjustments can cause climate drift over long periods of time within a model (Gordon et al., 2000).  If the ocean heat transports implied by atmospheric surface fluxes are significantly different from the ocean model, SST might drift to establish a new balance.  Thus, while flux adjustments might be useful for predicting ITCZ pattern, their universal effect could cause other undesirable issues.\\

Madden Julian oscillation is the dominant mode of tropical intraseasonal variability.  However, the IPCC reports that modelling of this phenomena remains challenging, leading to future projections of MJO having low confidence.  Like with precipitation change and ITCZ pattern, MJO has been shown to be highly sensitive to the pattern of SST warming (Maloney and Xie, 2012).  So, like before, the uncertainty in projecting future SST warming patterns is a leading cause of uncertainty in projecting MJO properties.\\

According to the IPCC report, most CMIP3 and CMIP5 models disagree on the response of the zonal SST gradient to the El Nino Southern Oscillation (ENSO) across the equatorial Pacific.  ENSO properties have been shown to be influenced by the background state of the tropical Pacific, which includes SST properties and the sensitivity of the atmosphere to SST changes (Yeh et al., 2012).  The magnitude of the SST trend in the tropical Pacific has been reduced from CMIP3 to CMIP5.  Yeh suggests that this change may be due to overestimation of the response to natural forcing and aerosols by including Earth system models.  If this is true, it emphasizes the need to examine how the addition of Earth System models, and their detailed physical processes, affects the larger models of which they are a part.\\

\subsection{Trends}

Within all these issues are a few common trends.  One trend is that many phenomena depend on the modelling of other phenomena.  And thus, if the modelling of one phenomena is flawed, than it would be unrealistic to expect accurate results from any phenomena that depend on it.  This is exemplified by the dependence of most tropical phenomena on the SST spatial and gradient patterns.  Since there are many questions raised about the accuracy of SST modelling, and the proper way to model (whether to use flux adjustments or not), this uncertainty extends to the modelling of precipitation patterns, ITCZ patterns, MJO, and ENSO.  While SST itself may not be very relevant to humans, the phenomena that depend on it certainly are.  Therefore, better modelling of SST is needed. \\

Another trend is the ‘trade-off’ between different modelling approaches.  Sometimes a particular approach is better at modeling a specific phenomenon than another approach.
However, this approach might have side-effects that negatively affect the model as a whole.
This is best exemplified by the tradeoff between using flux-adjustments or not.
As described above, using flux adjustments within a model can produce better results for simulating ITCZ, but as a consequence, can cause climate drift across the entire model as a whole.\\

The last trend is the importance of properly weighing the influence of various modelled processes.
If a particular process is excessively simulated its effects might exceed those seen in reality or diminish the effects of other important processes.  This is best exemplified by the overestimating of forcing that Yeh suggests is causing disparity in the SST trend of ENSO.
A key challenge in modelling is making sure some processes are not overly accounted for by multiple models.\\

Overall, the challenge for Applied Mathematicians moving forward is to better model SST in the tropics, or more specifically, better model the interaction between SST and atmosphere.  If the most advanced climate models had a more robust prediction of future SST warming, this by itself, would provide more consistency to predictions of precipitation, MJO, and ENSO.  It’s possible SST is not the only issue plaguing tropical climate predictions, but resolution of the issue would highlight the other issues which are not as noticeable at this time.\\

\subsection{References}

Brown, J., A. Moise, and R. Colman, 2012a: The South Pacific Convergence Zone in CMIP5 simulations of historical and future climate. Climate Dynamics, doi:10.1007/s00382-012-1591-x, 1-19.

Brown JR, Power SB, Delage FP, Colman RA, Moise AF, Murphy BF
(2011) Evaluation of the South Pacific Convergence Zone in
IPCC AR4 climate model simulations of the 20th century. J Clim
24:1565–1582

Gordon, C., Cooper, C., Senior, C., Banks, H., Gregory, J., Johns,
T., Mitchell, J., and Wood, R.: The simulation of SST, sea ice extents
and ocean heat transports in a version of the Hadley Centre
coupled model without flux adjustments, Climate Dynam., 16,
147–168, 2000.

Ma, J., and S.-P. Xie, 2013: Regional patterns of sea surface temperature change: A source of uncertainty in future projections of precipitation and atmospheric circulation. Journal of Climate, 26, 2482-2501.

Maloney, E. D., and S.-P. Xie, 2013: Sensitivity of MJO activity to the pattern of climate warming. Journal of Advances in Modeling Earth Systems, 5, 32-47.

Yeh, S.-W., Y.-G. Ham, and J.-Y. Lee, 2012: Changes in the tropical Pacific SST Trend from CMIP3 to CMIP5 and its implication of ENSO. Journal of Climate, 25, 7764-7771.

### 2013-11-03 project002/ipcc-morgan.tex
updating from class

\subsection{Summary}
\begin{itemize}
	\item it is believed tropical convection influences changes in SST (model)
	\item SPCZ GCMs and model do not correctly account for the orientation of SPCZ (model and data)
	\item ITCZ is modeled unrealistically in several prominent models (data)
	\item South Pacific island rainfall explained by ITCZ (model and data)
	\item SACZ explains flood/dry conditions  in south east Brazil (data) 
	\item ENSO may display variability explained by natural variability or by forcing. Models mimc observed data without external forcing, but better understanding of the underlying mechanisms is required to justify this assumption. 
	\item ENSO will remain a dominant mode of inter annual variability with global influences. Namely, ENSO induces rainfall intensity variability because of increased moisture 
\end{itemize}

\subsection{Wants}
\begin{itemize}
	\item there is a demand for improved modeling of SST warming patterns to improve rainfall estimates in climatological convergence zones
	\item improve annual estimates of SST warming patterns to the confidence of seasonal estimates
		\subitem{-} (COMPUTATIONAL SCIENTIST) improvements in prediction can be achieved with more computationally expensive model runs
	\item correct modeled orientation of SPCZ
		\subitem{-} (COMPUTATIONAL SCIENTIST) seems fair enough. I do not imagine this being computationally difficult
	\item more knowledge of zonally-oriented SPCZ events
	\item better knowledge of IOB formation from ENSO
	\item better understanding of annual SST change (data taken from coral isotopes)
	\item model for tropical Atlantic Climate variability in boreal summer
		\subitem{-} possibly incorporate feedback loops between surface wind, evaporation, and SST
	\item better predictions of AMM activity
	\item better MJO modeling 
		\subitem{-} sensitive to SST warming pattern, which itself has high uncertainty 
	\item demand for better understanding of how a warmer climate influences ENSO

\end{itemize}

### 2013-11-11 project003/UVM/finalProposal.tex
added project proposal

\documentclass[onecolumn]{article}
\title{Parameter Estimation in Toy Climate Models Utilizing a Genetic Algorithm}
\author{Morgan Frank \& Andy Reagan\\ Team UVM}
\usepackage{amsmath,amssymb}
\usepackage{mathptm,graphicx,rotate,color}
\hoffset=0in
\voffset=-.2in
\marginparwidth=0in
\marginparsep=0in
\oddsidemargin=0in
\topmargin=0in
\headheight=0in
\textwidth=470pt
\textheight = 620pt
\fontsize{12}{15}
\pagestyle{empty}

\begin{document}
\maketitle
\indent Suppose a scientist observes a climatological phenomenon and devises a model principled in understood climatological mechanisms.
The problem then becomes to find the parameters with which to run the model so as to best model the observed data.
Common parameter estimation approaches make use of regression analysis (e.g. shooting algorithms), Maximum Likelihood Estimation, Monte Carlo sampling, or simply empirical tuning.
We propose an alternative technique using a genetic algorithm for comparison.\\

\indent Genetic algorithms (GAs) borrow the ideas of reproduction, mutation, and fitness from biological evolution to search for real-valued vectors leading to optimal fitness.
Given a model and data, we will represent the parameters for the model as a real-valued vector and utilize Matlab's built-in GA implementation to attempt to find parameter choices that allow the model to best fit the data.
Our fitness function will be based on how well the model-run resulting from the parameters fits the data.
Initially, we intend to explore the effectiveness of GA on common toy models in dynamical systems and climate studies including the Lorenz '96 and '63 models, along with the Stommel model, Paillard model, and possibly other interesting models we discussed in class.
Executing the computation will require a model run for each fitness assessment; therefore it would be interesting to explore how useful and practical  GA for parameter fitting can be as we move up the model hierarchy.
Furthermore, we will explore the effectiveness of the GA as the dynamics of the toy models are changed to be more or less chaotic.
If we find that GA performs comparably to more commonly used techniques, then GA may be a useful alternative to scientists investigating climate modeling.
\end{document}

### 2013-11-24 project001/Paillard_1998/paillard_readme.tex
project 001 stuff

\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{D. Paillard's Conceptual Climate Model (\emph{Nature} 1998)}
\author{ }
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
\section{About the Models}
These models were created to explain the glacial-interglacial cycles within the framework of the orbital perturbations affecting incoming solar radiation, together with the feedbacks related to the Earth's ice sheets.

The first is a static model that changes states when thresholds are reached.  These thresholds are parameters in the model, and the states are \emph{i} (interglacial), \emph{g} (moderate glaciation), and \emph{G} (full glaciation).  The only transitions that are allowed are \emph{i}-\emph{g}, \emph{g-G} and \emph{G-i}, and all of these transitions take place when the insolation (read in at model run time) satisfies certain conditions related to each threshold.  See the figure below for a schematic diagram of the transitions that take place.

The second model is continuous in the extent of the ice sheet volume, and is otherwise just like the first static model. There are two time-related parameters in this model which can be tuned to match the data.

The insolation is provided by the file \texttt{ins\_65N\_July.txt}, which is the average insolation for the month of July at a latitude of 65 degrees north and is taken from Berger et al (1991).  The plots of the oxygen isotope stacks, which are a proxy for surface temperature, are taken from Bassinot et al (1994).

\section{Running the Model}
To run the models without changing parameters, simply type \texttt{paillard\_steady\_run} at the command line for the static model, or \texttt{paillard\_conts\_run} for the continuous time model.  The model will display three panels:  the insolation, normalized to mean 0 and unit variance, the ice volume, and the oxygen isotope stack.  To change parameters, edit the \texttt{paillard\_parameters.m} file, which has a description of each parameter in the comments at the beginning of the file.

Also included are the insolation file (\texttt{ins\_65N\_July.txt}) and the oxygen isotope stack file (\texttt{bassinot.txt}), but these should be considered as data and not altered.  The model is attempting to use the insolation data to simulate the oxygen isotope stack data.

\section{Reference}
\noindent Paillard, D. "The Timing of Pleistocene glaciations from a simple multiple-state climate model." \emph{Nature}.  Vol 391.  22-Jan-1998.  pp 378-381.\\
\\
\noindent Berger, A. and Loutre M.F.  "Insolation values for the climate of the last 10 million years."  \emph{Quaternary Sciences Review}, Vol 10, No. 4, pp. 297-317, 1991.\\
\\
\noindent Bassinot, FC et al. (1994): A low-latitude Upper Pleistocene oxygen isotope stack, doi:10.1594/PANGAEA.715006.\\
\indent Supplement to: Bassinot, Frank C; Labeyrie, Laurent; Vincent, Edith; Quidelleur, Xavier; Shackleton, Nicholas J; Lancelot, Yves (1994): The astronomical theory of climate and the age of the Brunhes-Matuyama magnetic reversal, Earth and Planetary Science Letters, 126(1-3), 91-108, doi:10.1016/0012-821X(94)90244-5
\\ \\ \\ \\ \\
\begin{figure}[h]
\includegraphics[scale=1.0]{paillard_model_cartoon.pdf}
\caption{Schematic of Model States, Transitions and Thresholds from Paillard (1998)}
\end{figure}
\end{document}

### 2013-11-24 project001/Paillard_1998/isotope_read.m

@@ -1,0 +1,7 @@
+%Read in Oxygen Isotope Data
+
+A = dlmread('bassinot.txt');
+
+isotope_time = A(:,1);
+isotope = A(:,2);
+

### 2013-11-24 project001/Paillard_1998/paillard_conts_run.m

@@ -1,0 +1,131 @@
+% Implementation of Continous State Model in 
+% Paillard's 1998 Letter to Nature:
+% Paillard,D. "The Timing of Pleistocene Glaciations from a Simple Multiple
+% State Climate Model." Nature. 391. 1/22/1998.
+%
+% The model allows an ice volume to vary continuously between regimes:
+% i = interglacial
+% g = moderate glaciation
+% G = full glaciation
+%
+% according to the ordinary differential equation
+%
+% dv/dt = (v_R-v)/tau_R - F/tau_F
+%
+% where R is i, g, or G, and F is the "truncated" insolation taken from
+% Bergers data at 65 deg N in July (NCDC, 1992) and truncated by 
+% f(x)=x+sqrt(a+x^2).
+%
+% The ice regime is determined by three thresholds which give the following
+% transitions:
+% i-g: insolation drops below i_0
+% g-G: ice volume reaches v_max
+% G-i: insolation exceeds i_1
+%
+% The states are 
+% i = 0
+% g = 0.5
+% G = 1
+%
+% The results are compared against an isotope record detailed in Bassinot
+% et al (NCDC, 1994) (Bassinot.txt).
+%------------------------------------------------------------------------
+
+clear
+
+%Initialize Arrays
+v=zeros(900,1);
+state = zeros(900,1);
+
+%Load Parameter File
+
+paillard_parameters
+
+%Read in insolation data and normalize it to zero mean, unit variance
+
+A = dlmread('ins_65N_July.txt');
+
+%Use only last 1 million years
+
+time = A(1:901,1);
+insolation = A(1:901,2);
+
+%Truncate Using Truncation Function
+
+insolation = truncation_f(insolation,a);
+
+%Subtract Mean and normalize Var to 1
+
+insolation = (insolation-mean(insolation))/sqrt(var(insolation));
+
+%Initial Condition for Ice Regimes
+state(1)=1;
+v(1)=0.75;
+for i=2:901
+    switch state(i-1)
+        case(0)
+             if (insolation(i)<i_0)
+                state(i)=1;
+                v_current = v_g;
+                tau_current = tau_g;
+            else
+                state(i)=0;
+                v_current = v_i;
+                tau_current = tau_i;
+             end
+        case(1)
+            if (v(i-1)>=v_max)
+                state(i)=2;
+                v_current = v_capg;
[…]

### 2013-11-24 project001/Paillard_1998/paillard_parameters.m

@@ -1,0 +1,42 @@
+%Parameters for Models from Paillard's 1998 Letter to Nature:
+%Paillard,D. "The Timing of Pleistocene Glaciations from a Simple Multiple
+%State Climate Model." Nature. 391. 1/22/1998.
+%
+%   i_0: initiates i-g transition
+%   i_1: initiates G-i transition
+%   i_2: initiates g-G transition
+%   i_3: constrains g-G transition (only if insolation remains < i_3)
+%   t_g: time required for ice volume to grow to v_max and initiate g-G
+%   tau_capg: time scale for state G
+%   tau_g: time scale for state g
+%   tau_i: time scale for state i   
+%   tau_F: time scale for smoothed insolation forcing
+%   v_max: sufficient ice volume to trigger g-G (implied)
+%   v_capg: ice volume in state G
+%   v_g: ice volume in state g
+%   v_i: ice volume in state i
+%   a: parameter for insolation truncation
+%
+%-------------------------------------------------------------------------
+
+%Insolations are in terms of variances
+i_0 = -0.75;
+i_1 = 0.;
+i_2 = 0.;
+i_3 = 1.;
+
+%Ice volumes (normalized)
+v_capg = 1;
+v_g = 1;
+v_i = 0;
+v_max = 1;
+
+%Time Constants
+t_g = 33; %kyr
+tau_capg = 50;
+tau_g = 50;
+tau_i = 10;
+tau_F = 25;
+
+%Insolation truncation parameter
+a=1;

### 2013-11-24 project001/Paillard_1998/paillard_steady_run.m

@@ -1,0 +1,120 @@
+%Implementation of Steady Model in 
+%Paillard's 1998 Letter to Nature:
+%Paillard,D. "The Timing of Pleistocene Glaciations from a Simple Multiple
+%State Climate Model." Nature. 391. 1/22/1998.
+%
+%The model allows three discrete states:
+%i = interglacial
+%g = moderate glaciation
+%G = full glaciation
+%
+%with transitions being triggered by:
+%i-g: insolation drops below threshold i_0
+%g-G: ice volume exceeds threshold v_max
+%G-i: insolation exceeds threshold i_1
+%
+%Note that no other transitions are allowed, and that the g-G transition
+%only occurs if insolation remains below some maximum i_3 for some
+%specified time t_g, and the transition occurs when insolation drops below
+%i_2.
+%
+%Thus the model has parameter set i_0, i_1, i_2, i_3, t_g, and v_max.
+%
+%The results are compared to the isotope record pulled from ice cores,
+%seeking agreement with major ice cover shifts from the record.
+%
+%Array "state" has entries that consist of
+%  2 if state = i
+%  1 if state = g
+%  0 if state = G
+%------------------------------------------------------------------------
+
+clear
+
+%Read in insolation data and normalize it to zero mean, unit variance
+
+A = dlmread('ins_65N_July.txt');
+
+%Use only last 1 million years
+
+time = A(1:901,1);
+insolation = A(1:901,2);
+
+%Subtract Mean and normalize Var to 1
+
+insolation = (insolation-mean(insolation))/sqrt(var(insolation));
+
+
+%Initialize arrays
+state=zeros(size(time));
+state=state-9999;   %just to debug code
+
+%Load Parameter File
+paillard_parameters
+
+%Initialize Model
+state(1)=0;
+
+for i=2:901
+    switch state(i-1)
+        case(0)
+            if (insolation(i)>i_1)
+                state(i)=2;
+            else
+                state(i)=0;
+            end
+        case(1)
+            check=0;
+            if (i>34)
+                for j=i-34:i-1
+                    if ((state(j)==1) && (insolation(j)<=i_3))
+                        check = check;
+                    else
+                        check = check+1;
+                    end
+                end
+                if ((check==0) && (insolation(i)<i_2))
+                    state(i)=0;
+                else
[…]

### 2013-11-24 project001/Paillard_1998/truncation_f.m

@@ -1,0 +1,5 @@
+function z=truncation_f(x,a)
+
+z = x.^2+sqrt(a+x.^2);
+
+end

### 2013-11-24 project003/UVM/DA-code/lorenz63_paramEst.m
project 003 materials

@@ -1,0 +1,156 @@
+classdef lorenz63_paramEst<handle
+	%lorenz63: class for a lorenz63 run
+	%   Goal is to implement a class for which MATLAB can
+	%   control and run the lorenz63 model
+	%
+	%   Additionally, this is a 6 dimensional system, including params
+	%
+	%   USAGE:
+	%       -initialize the model with a climatological IC
+	%        ens1 = lorenz63();
+	%        ens1.init();
+	%        OR with a name
+	%        ens1.init('name')
+	%        OR initialize with given IC (and name)
+	%        ens1.init('name',[10.4;11.7;18.5]);
+	%       -set the IC for the run to something specific
+	%        ens1.x = [10.4;11.7;18.5];
+	%       -run this case
+	%        ens1.run();
+	%       -run more into the future
+	%        ens1.run();
+	
+	properties
+		%basics
+		x
+		tstep = .01;
+		dim = 6;
+		time = 0;
+		window = 20;
+		
+		params = {};
+		%for the TLM
+		%this code only works for rk2prime...
+		TLMmethod = 'rk2prime'; % 'kam','explicit'
+		p_f
+		DIR
+	end % properties
+	methods
+		function self = lorenz63(varargin)
+			%intialize the class
+		end %constructor
+		function init(self,varargin)
+			self.x = 10*randn(self.dim,1);
+			self.x(4:6) = [10 0 0;0 20 0; 0 0 40]*rand(3,1); %3*randn(3,1)+[8/3;10;28];
+			%fprintf('initializing...');
+			
+			% ignore the first argument, if there is one
+			if nargin > 2
+				% setting given IC
+				self.x = varargin{2};
+			else
+				% generating climatological IC
+				tmpwindow = self.window; self.window = randi(100,1);
+				tmptime = self.time; self.time = 0;
+				self.x(1:3) = [-12.6480;-13.9758;30.9758]+.1*randn(3,1);
+				self.run();
+				self.window = tmpwindow;
+				self.time = tmptime;
+			end
+			%fprintf('done\n');
+		end %init
+		function run(self,varargin)
+			% run the case
+			%fprintf('running...');
+			
[…]

### 2013-11-24 project003/UVM/DA-code/run_filter.m

@@ -1,0 +1,148 @@
+% run_filter.m
+%
+% test parameter estimation of data assimilation
+%
+% Andy Reagan
+% 2013-11-23
+
+clear all
+close all
+addpath(genpath('/Users/andyreagan/work/2013/2013-05data-assimilation/src'))
+
+%% fixed
+dim = 6;
+
+%% tunable parameters
+numRuns = 1;
+obs_error_std = 0.05;
+runTime = 50; %  model units
+window = .5; %  window
+modelname = 'lorenz63';
+params = {8/3,10,28};
+obsVar = 1:3;
+
+%% build observation operator
+H = zeros(length(obsVar),dim);
+H(obsVar,obsVar) = eye(length(obsVar));
+R = obs_error_std*eye(length(obsVar));
+
+%% initialize storage
+num_windows = ceil(runTime/window);
+truth_vec = repmat(zeros(dim,num_windows+1),1,numRuns);
+
+%%  make a truth run
+truthmodel = lorenz63();
+truthmodel.params = params;
+truthmodel.init();
+truthmodel.window = window;
+% save the IC because I've give the DA,noDA runs good IC?
+tmp_truth_vec(1:3,1) = truthmodel.x;
+% generate the truth x timeseries
+for j=1:num_windows
+	truthmodel.run()
+	tmp_truth_vec(1:3,j+1) = truthmodel.x;
+end
+truth_vec(1:3,1) = tmp_truth_vec(1:3,1);
+truth_vec(1:3,2:end) = repmat(tmp_truth_vec(:,2:end),1,numRuns);
+% fill in the truth parameters
+truth_vec(4:6,:) = repmat([params{1};params{2};params{3}],1,length(truth_vec(1,:)));
+
+% pass all of the observations, including time 0
+obs_pert = truth_vec(obsVar,:) + obs_error_std*randn(size(truth_vec(obsVar,:)));
+[EKF_f_vec,EKF_a_vec] = modelDAinterface(@lorenz63_paramEst,'EKF',obs_pert,0:window:(runTime*numRuns),H,R,window,0,0);
+[ETKF_f_vec,ETKF_a_vec] = modelDAinterface(@lorenz63_paramEst,'ETKF',obs_pert,0:window:(runTime*numRuns),H,R,window,0,0); 
[…]

### 2013-11-26 project003/UVM/DA-code/runDAParamFitAll.sh
setting up vacc param run

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+# need these for compiled matlabs jobs
+export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/gpfs1/arch/x86_64/matlab2010a/bin/glnxa64:/gpfs1/arch/x86_64/matlab2010a/runtime/glnxa64:/gpfs1/arch/x86_64/matlab2010a/sys/os/glnxa64:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/native_threads:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/server:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/client:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64
+export XAPPLRESDIR=/gpfs1/arch/x86_64/matlab2010a/X11/app-defaults
+
+cd /users/a/r/areagan/work/2013/data-assimilation/src/experiments/openFoamAssimilation
+
+# C=40
+# SEED=0
+# for B in 2 4 8 16 32 
+# do
+#   for D in 10 20 30
+#   do
+#     for E in 2 3 5 10 20
+#     do
+#       echo $B $C $D $E
+#       qsub -v NUMTSENSORS="$B",NUMWINDOWS="$C",WINDOWLEN="$D",NUMENS="$E",COUNTER="$SEED" runFullModelLocalZones.qsub
+#       SEED=$(($SEED+1))
+#     done
+#   done
+# done
+
+## for a single test (always find something wrong)
+C=40
+D=10
+E=20
+SEED=100
+for B in 32 100 1000 2000 5000 10000 20000 40000
+do
+  qsub -v NUMTSENSORS="$B",NUMWINDOWS="$C",WINDOWLEN="$D",NUMENS="$E",COUNTER="$SEED" run.qsub
+done
+
+
+

### 2013-11-26 project003/UVM/DA-code/run_filter.m

@@ -8,6 +8,7 @@
 clear all
 close all
 addpath(genpath('/Users/andyreagan/work/2013/2013-05data-assimilation/src'))
+addpath(genpath('/users/a/r/areagan/work/2013/data-assimilation/src'))
 
 %% fixed
 dim = 6;

### 2013-12-09 project003/UVM/paper/mcrn501-p3.body.tex
set up for final paper

--- 2013-09-17 project001/UVM/paper/mcrn501-p1-uvm.body.tex
@@
-\section{Introduction}
+section{Introduction}
 
@@
 \centering
-  \includegraphics[width=.48\textwidth]{../data/dO18/d18O_lisiecki_noname.pdf}
+  \includegraphics[width=.48\textwidth]{figures/d18O_lisiecki_noname.pdf}
   \caption{
@@
 \centering
-  \includegraphics[width=.45\textwidth]{../src/discrete_plot_noname.pdf}
+  \includegraphics[width=.45\textwidth]{figures/discrete_plot_noname.pdf}
   \caption{
@@
 \centering
-  \includegraphics[width=.48\textwidth]{../data/insolation/laskar2004/insol_data2_noname.pdf}
+  \includegraphics[width=.48\textwidth]{figures/insol_data2_noname.pdf}
   \caption{
@@
 \centering
-  \includegraphics[width=.48\textwidth]{cmaes.png}
+  \includegraphics[width=.48\textwidth]{figures/cmaes.png}
   \caption{
@@
 \centering
-  \includegraphics[width=.48\textwidth]{gaImg.png}
+  \includegraphics[width=.48\textwidth]{figures/gaImg.png}
   \caption{
@@
 \centering
-  \includegraphics[width=.48\textwidth]{../src/results_plot_noname.pdf}
+  \includegraphics[width=.48\textwidth]{figures/results_plot_noname.pdf}
   \caption{

### 2013-12-09 project003/UVM/paper/makerevtex4.pl

@@ -1,0 +1,103 @@
+#!/usr/bin/perl
+
+foreach $switch (@ARGV) {
+    if (($switch eq "-quick") or ($switch eq "-q"))
+    {
+	$quickswitch = 1;
+    }
+}
+
+$bodylist = `ls *.body.tex`;
+@bodies = split(/\s+/,$bodylist);
+
+($sec,$min,$hour,$numday,$month,$year,$wday,$yday,$isdst) = localtime(time);
+$year += 1900;
+$timestamp = "$year $month $numday $hour $min $sec";
+
+foreach $body (@bodies) {
+    ($filename = $body) =~ s/\.body\.tex$//;
+    if (-e "$filename-revtex4.tex") {
+	($logfilename = "log/progress-$body") =~ s/\.body\.tex$/-log/;
+	print "processing $filename...\n";
+
+	# check for postscript
+	open (BODY, "$body") or die "can't open $body: $!\n";
+	undef $/;
+	$bodytex = <BODY>;
+	close BODY;
+	if ($bodytex =~ m/\.ps\}/msg) {
+	    $pdfswitch = 0;
+	}
+	else {
+	    $pdfswitch = 1;
+	}
+	    
+	if ($pdfswitch == 1) {
+	    `pdflatex $filename-revtex4 1>&2`;
+	}
+	else {
+	    `latex $filename-revtex4 1>&2`;
+	}
+	unless ($quickswitch == 1) {
+	    `bibtex $filename-revtex4 1>&2`;
+	    if ($pdfswitch == 1) {
+		`pdflatex $filename-revtex4 1>&2`;
+		`pdflatex $filename-revtex4 1>&2`;
+	    }
+	    else {
+		`latex $filename-revtex4 1>&2`;
+		`latex $filename-revtex4 1>&2`;
+	    }
+	}
+	if ($pdfswitch == 0) {
+	    `dvips -o $filename-revtex4.ps $filename-revtex4 1>&2`;
+	    `epstopdf $filename-revtex4.ps 1>&2`;
+	}
+
+	# if open, bring to front
+	# `open $filename-revtex4.pdf`;
+	`open -a preview`;
+
+	`countcheckboxes $filename.abs.tex $filename.body.tex $filename.supplementary.tex 1>&2`;
+
+	$numtodos = `countcheckboxes-total $filename.abs.tex $filename.body.tex`;
+	chomp $numtodos;
+
+	# extract page and byte numbers
[…]

### 2013-12-09 project003/UVM/paper/setup.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+#
+# set up these files for YOUR disseration!
+#
+# USAGE:
+# ./setup.sh reagan-thesis
+
+OLD_NAMEBASE="mcrn501-p1-uvm"
+NEW_NAMEBASE=$1
+
+for FILE in *.tex
+do
+  mv $FILE $(echo $FILE | sed s/$OLD_NAMEBASE/$NEW_NAMEBASE/)
+done
+
+\rm $OLD_NAMEBASE*
+
+sed -i .old s/$OLD_NAMEBASE/$NEW_NAMEBASE/ $NEW_NAMEBASE-revtex4.settings.tex
+
+\rm *.old

### 2013-12-10 project003/UVM/finalPaper/paper.tex
working on final paper

\documentclass[onecolumn]{article}
\usepackage{cite}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{abstract}
 \usepackage{fullpage}
 \usepackage{rotating}
 \usepackage{multicol}
 \usepackage{multirow}
 \usepackage{array}
 \usepackage[abs]{overpic}
 \setlength\unitlength{1mm}
 \usepackage{mathptm,graphicx,rotate,color}
% \title{Spectral Bifurcation Diagrams Reveal Order for many Spatial Dimensions in the Lorenz '96 system\\ Spatial Dimension Exploration leads to Stable Standing Waves in the Lorenz '96 Model\\ }
 \title{Data Assimilation and Genetic Algorithms for the Parameter Estimation Problem in Simple Climate Models}
 \author{Morgan R. Frank$^{1}$, Andrew Reagan$^{2}$\\
 {\footnotesize  Computational Story Lab, Department of Mathematics and Statistics, Vermont Complex Systems Center, Vermont Advanced Computing Core,}\\
{\footnotesize University of Vermont, Burlington, Vermont, United States of America }\\
{\footnotesize$^{1}$mrfrank@uvm.edu, $^{2}$andrew.reagan@uvm.edu}}

\begin{document}
\maketitle
\begin{multicols}{2}
\begin{abstract}
	\bf\indent Given observations of an atmospheric phenomenon and a well-principled model of that phenomenon, the parameters for the model must be properly tuned if the model is to mimic the data. We investigate the use of genetic algorithm in comparison to data assimilation as a means of performing parameter estimation when tuning models to data. We compare results while tuning chaotic dynamics, observation noise and frequency, and system dimensionality while performing parameter estimation for the Lorenz '63 and Lorenz '96 systems.
\end{abstract}
\section{Introduction}
\indent Weather forecasts have become an expected part of everyday life in the modern society. Things like air-travel, disaster preparation, and daily planning rely on effective predictions \cite{kerr}. However, predicting future states of the atmosphere proves to be difficult as chaotic systems exhibit sensitive dependence of initial conditions \cite{farmer,orrell3,D+Y}. This hurdle is overcome by utilizing computationally expensive global climate models (GCMs), but scientists working to improve weather forecasting often lack the time or computational power to execute many GCMs. Instead, climate scientists often use simple models that account for particular aspects of the weather forecasting problem. \\
\indent Edward Lorenz has made major contributions to the fields of dynamical systems and atmospheric prediction \cite{lorenz95,lorenz68,lorenz98}. Two such contributions are the wildly popular Lorenz '63 system \cite{lorenzAttr} and the Lorenz '96 system \cite{lorenz96}. The Lorenz '63 system (L63), which yields the widely known Lorenz Attractor, is a simple three-variable model with highly tunable dynamics, allowing researchers a computationally tractable means to experiment in the predictability of chaotic systems. The Lorenz '96 system (L96) exhibits tunable chaotic dynamics as well, while additionally providing a computationally tractable way to change the system dimensions and tune the accuracy of data observations. Both systems provide interesting and computationally manageable test beds for the parameter estimation problem across several different types of systems. Figure 1 shows example trajectories for each system.\\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
	\subsection{The Lorenz '63 Model}

	\subsection{The Lorenz '96 Model}
\indent In 1995, Edward Lorenz introduced the following $I$-dimensional model \cite{lorenz95,lorenz98}. The key characteristics of this model include tunable chaotic behavior when subject to enough forcing, and tunable dimensionality. The predecessor to the current model is given by
 \begin{equation}
 	\frac{dx_{i}}{dt}=x_{i-1}(x_{i+1}-x_{i-2})-x_{i}+F
 \end{equation}
 where $i=1,2,\dots,I$ and $F$ is the forcing parameter. Each $x_{i}$ represents observations of some atmospheric atmospheric quantity, like temperature, evenly distributed about a given latitude of the globe. This implies a modularity in the indexing that is described by $x_{i+I}=x_{i-I}=x_{i}$. \\
 \indent This early model failed to produce realistic growth rate of the large-scale errors along with lacking tenability in observation reliability. Lorenz went on to introduce 
\end{multicols}
\begin{center}
	$\begin{array}{cc}
		\begin{overpic}[width=.5\columnwidth,trim=1cm 6cm 1cm 7cm,clip]{L63Traj}\put(20,40){\colorbox{white}{\fbox{A}}}\end{overpic}&
		\begin{overpic}[width=.5\columnwidth,trim=3cm 7cm 1cm 6cm,clip]{chaotic_30_5_14}\put(20,40){\colorbox{white}{\fbox{B}}}\end{overpic}\\
		\begin{overpic}[scale=.4,trim=1cm 5cm 0cm 7cm,clip]{L63_X_exp}\put(15,55){\colorbox{white}{\fbox{C}}}\end{overpic}&
		\begin{overpic}[scale=.5,trim=3cm 7cm 0cm 8cm,clip]{L96_slow_exp}\put(20,55){\colorbox{white}{\fbox{D}}}\end{overpic}\\
%		\begin{overpic}\end{overpic}
%		\begin{overpic}\end{overpic}
	\end{array}$
\end{center}
{\bf\small Figure 1. (A) The popular ``Lorenz Attractor" produced with the Lorenz '63 system. This three-variable system produces a ``butterfly"-like chaotic attractor that is well-known among fractal and chaos enthusiasts. (B) An snapshot of a trajectory of the Lorenz '96 system. Each blue point is a slow oscillator, and the adjacent sections of green represent the fast oscillators coupled with the corresponding slow oscillator. The origin represents the lowest value achieved by any of the slow oscillators on this trajectory. The red line is a cubic spline interpolation of the blue data points. (C) An example trajectory of the $X$ variable from the Lorenz '63 system. (D) An example trajectory for a slow oscillator of the Lorenz '96 system.}\\
%Consider adding single variable trajectories.
\begin{multicols}{2}
a more flexible model in 1996 by coupling two systems similar to the model in equation (1), but differing in time scales. The equations for the Lorenz '96 model \cite{lorenz96} are given as
 \begin{equation}
 	\frac{dx_{i}}{dt}=x_{i-1}(x_{i+1}-x_{i-2})-x_{i}+F-\frac{hc}{b}\displaystyle\sum_{j=1}^{J}y_{(j,i)}
 \end{equation}\vspace{-1cm}
 \begin{equation}
 	\frac{dy_{(j,i)}}{dt}=cby_{(j+1,i)}(y_{(j-1,i)}-y_{(j+2,i)})-cy_{(j,i)}+\frac{hc}{b}x_{i}
 \end{equation}
 where $i=1,2,\dots,I$ and $j=1,2,\dots,J$. The parameters $b$ and $c$ indicate the time scale of solutions to equation (3) relative to solutions of equation (2), and $h$ is the coupling parameter. The coupling term can be thought of as a parameterization of dynamics occurring at a  spatial and temporal scale unresolved by the $x$ variables. Again, each $x_{i}$ represents an atmospheric observation about a latitude that oscillates in slow time, and the set of $y_{(j,i)}$ are a set of $J$ fast time oscillators that act as a damping force on $x_{i}$. The $y$'s exhibit a similar modularity described by $y_{(j+IJ,i)}=y_{(j-IJ,i)}=y_{(j,i)}$.

 	\subsection{Data Assimilation}

	\subsection{Genetic Algorithm}
\indent 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{9}
	\bibitem{kerr}
	R. A. Kerr. \emph{Weather Forecasts Slowly Clearing Up}. Science 9 November 2012: Vol. 338 no. 6108 pp. 734-737 DOI: 10.1126/science.338.6108.734
        \bibitem{farmer}
                Farmer, J. D., and J. J. Sidorowich. \emph{Predicting Chaotic Time Series.} Phys. Rev. Lett. 59(8) (1987): 845-848.
	\bibitem{orrell3}
	 D. Orrell, \emph{Role of the Metric in Forecast Error Growth: How Chaotic is
the Weather?}, Tellus 54A (2002) 350Ð362.
        \bibitem{D+Y}
        	C. M. Danforth, J. A. Yorke.  2006. \emph{Making Forecasts for Chaotic Physical Processes.}
		Physical Review Letters, 96, 144102.
 	\bibitem{lorenz95}
	E.N. Lorenz, \emph{Predictability A problem partly solved}, in: ECMWF Seminar Proceedings on Predictability, Reading, United
Kingdom, ECMWF, 1995, pp. 118.
         \bibitem{lorenz68}
                E. N. Lorenz. \emph{The predictability of a flow which possesses many scales of motion}. {\bf Tellus XXI}, 289 (1968).
	\bibitem{lorenz98}
	E.N. Lorenz, K.A. Emanuel, \emph{Optimal sites for supplementary weather observations: simulation with a small model}, J. Atmos. Sci.
55 (1998) 399Ð414.
         \bibitem{lorenzAttr}
                Lorenz, Edward N., 1963: \emph{Deterministic Nonperiodic Flow}. {\it J. Atmos. Sci.}, {\bf 20}, 130.141.
           \bibitem{lorenz96} 
                E. N. Lorenz, Proc. Seminar on Predictability 1, 1 (1996).
        \bibitem{england}
                R. England. \emph{Error estimates for Runga-Kutta type solutions to systems of ordinary differential equations}. {\it The Computer Journal} (1969) 12 (2): 166-170. doi: 10.1093/comjnl/12.2.166
%        \bibitem{werneth}
%                Charles M Werneth et al 2010 Eur. J. Phys. {\bf (31)} 693 doi:10.1088/0143-0807/31/3/027
        \bibitem{wilks}
         D. S. Wilks, \emph{Effects of Stochastic Parametrizations in the Lorenz Õ96 System}, Quart. J. Roy. Meteo. Soc. 131 (2005) 389Ð407.
\end{thebibliography}
\end{multicols}
\end{document}

### 2013-12-10 project003/UVM/finalPaper/paper.tex
made my updates

@@
 \usepackage{abstract}
+\usepackage{tabularx}
 \usepackage{fullpage}
@@
 
+\allowdisplaybreaks
+\widowpenalty=1000000
+\clubpenalty=1000000
+
 \begin{document}
@@
 \indent Weather forecasts have become an expected part of everyday life in the modern society.
-Things like air-travel, disaster preparation, and daily planning rely on effective predictions \cite{kerr}.
-However, predicting future states of the atmosphere proves to be difficult as chaotic systems exhibit sensitive dependence of initial conditions \cite{farmer,orrell3,D+Y}.
-This hurdle is overcome by utilizing computationally expensive global climate models (GCMs), but scientists working to improve weather forecasting often lack the time or computational power to execute many GCMs.
-Instead, climate scientists often use simple models that account for particular aspects of the weather forecasting problem.
-\\
+Things like air-travel, disaster preparation, and daily planning rely on accurate predictions \cite{kerr}.
+However, predicting future states of the atmosphere proves to be difficult, as chaotic systems exhibit sensitive dependence of initial conditions, and the underlying processes in weather are  known to be chaotic \cite{farmer,orrell3,D+Y}.
+This hurdle is overcome by utilizing computationally expensive global forecasting systems (GFS) for prediction and advanced methods for initial condition determination, but scientists working to improve weather forecasting often lack the time or computational power to execute many high resolution GFS experiments.
+Instead, climate scientists often use simple models that account for particular aspects of the weather forecasting problem.\\
 \indent Edward Lorenz has made major contributions to the fields of dynamical systems and atmospheric prediction \cite{lorenz95,lorenz68,lorenz98}.
@@
 \subsection{The Lorenz '63 Model}
+In 1962, Barry Saltzmann attempted to model convection in a Rayleigh-B\'{e}rnard cell  by reducing the equations of motion into their core processes \cite{saltzman1962finite}.
+Then in 1963 Edward Lorenz reduced this system ever further to 3 equations, leading to his landmark discovery of deterministic non-periodic flow \cite{lorenz1963}.
+This system, which we will call the Lorenz 63 system, exhibits sensitive dependence on initial conditions, meaning that small errors in an approximation will lead to exponential error growth.
+These equations have since been the subject of intense study and have changed the way we view prediction and determinism, remaining the simple system of choice for examining nonlinear behaivor today \cite{kalnay20074}.
+The three equations are:
+\begin{align*}
+\frac{dx}{dt} &= \sigma (y-x)\\
+\frac{dy}{dt} &= \rho x - y -xz \\
+\frac{dz}{dt} &= xy -  \beta z .\end{align*}
 
-	\subsection{The Lorenz '96 Model}
-\indent In 1995, Edward Lorenz introduced the following $I$-dimensional model \cite{lorenz95,lorenz98}.
-The key characteristics of this model include tunable chaotic behavior when subject to enough forcing, and tunable dimensionality.
-The predecessor to the current model is given by
- \begin{equation}
- 	\frac{dx_{i}}{dt}=x_{i-1}(x_{i+1}-x_{i-2})-x_{i}+F
- \end{equation}
- where $i=1,2,\dots,I$ and $F$ is the forcing parameter.
-Each $x_{i}$ represents observations of some atmospheric atmospheric quantity, like temperature, evenly distributed about a given latitude of the globe.
-This implies a modularity in the indexing that is described by $x_{i+I}=x_{i-I}=x_{i}$.
-\\
- \indent This early model failed to produce realistic growth rate of the large-scale errors along with lacking tenability in observation reliability. Lorenz went on to introduce 
+The cannonical choice of $\sigma = 10, \beta = 8/3$ and $\rho = 28$ produce the well known butterfly attractor, and to adjust the strength of nonlinearity (chaos) we tune the $\rho$ parameter.
+
 \end{multicols}
@@
 \begin{multicols}{2}
+
+\end{multicols}
+\begin{center}
+\begin{tabular}{lll}
+  \hline
+  Parameter & Values Explored & Interpretation, if any\\
+  \hline
+  \hline
+  Observed Variables (63 Only) & $[x_1,$all] & Limited observations\\
+  \hline
+  Observational Noise & Normal in [0,.01,.05,.1,.25,.5,1,2] & Measurement and representativeness errors\\
+  ~~~~~~~~~~~~~~~~~~~ & Uniform in [0,.5,2,4,6,8,10] & \\
+  \hline
+  Nonlinearity & $\rho$ in [22,28,35] or $I$ in [4,8,10,15] & Chaotic behavior\\
+  \hline
+  Subsampled observations & [1,5,25,50] & Infrequent observations\\
+  \hline
+\end{tabular}
+\vspace{3mm}
+
+{\bf\small Table 1: Experimental parameter choices on which we test the performance of Data Assimilation and a Genetic Algorithm for fitting model parameters.}
+\end{center}
+\begin{multicols}{2}
+
+\subsection{The Lorenz '96 Model}
+\indent In 1995, Edward Lorenz introduced the following $I$-dimensional model \cite{lorenz95,lorenz98}.
+The key characteristics of this model include tunable chaotic behavior when subject to enough forcing, and tunable dimensionality.
+The predecessor to the current model is given by
+\begin{equation}
+\frac{dx_{i}}{dt}=x_{i-1}(x_{i+1}-x_{i-2})-x_{i}+F
+\end{equation}
+where $i=1,2,\dots,I$ and $F$ is the forcing parameter.
+Each $x_{i}$ represents observations of some atmospheric atmospheric quantity, like temperature, evenly distributed about a given latitude of the globe.
+This implies a modularity in the indexing that is described by $x_{i+I}=x_{i-I}=x_{i}$. \\
+\indent This early model failed to produce realistic growth rate of the large-scale errors along with lacking tenability in observation reliability.
+Lorenz went on to introduce
+
 a more flexible model in 1996 by coupling two systems similar to the model in equation (1), but differing in time scales. The equations for the Lorenz '96 model \cite{lorenz96} are given as
@@
 
+Areas as disparate as quadcopter stabilization \cite{achtelik2009visual} to the tracking of ballistic missle re-entry \cite{siouris1997tracking} use data assimilation.
+The purpose of data assimilation in weather prediction is defined by Talagrand as ``using all the available information, to determine as accurately as possible the state of the atmospheric (or oceanic) flow.'' \cite{talagrand1997assimilation}
+The data assimilation algorithm that we use here, the Kalman filter, was originally implemented in the navigation system of Apollo program \cite{kalman1961new,savely1972}.
+
+Data assimilation algorithms consist of a 3-part cycle: predict, observe, and assimilate.
+Formally, the data assimilation problem is solved by minimizing the initial condition error in the presence of specific constraints.
+The prediction step involves making a prediction of the future state of the system, as well as the error of the model, in some capacity.
+Observing systems come in many flavors: rawindsomes and satellite irradiance for the atmosphere, temperature and velocity reconstruction from sensors in experiments, and sampling the market in finance.
+Assimilation is the combination of these observations and the predictive model in such a way that minimizes the error of the initial condition state, which we denote the analysis.
+
+In addition to determining the initial conditions, we can extend the Extended Kalman Filter (EKF) to determine the model parameters.
+This is accomplised by considering the model parameters as variables of the model itself, with their differential equation being equal to 0, since they do not change with the solution.
+The value of this consideration is that the covariance of the model variables and model parameters is now included in the Tangent Linear Model (the Jacobian of the extended analytical system) and hence is updated by the Kalman gain matrix.
+
+The formulation of the filter we employ is the standard formulation, since the incorporation of parameters into the estimation is independent of the filter itself.
+Using the notation of Kalnay \cite{kalnay2003}, this amounts to making a forecast with the nonlinear model $M$ (either Lorenz 63 or Lorenz 96 in this study), and updating the error covariance matrix $\mathbf{P}$ with the TLM $L$, and adjoint model $L^T$
+\begin{align*} \mathbf{x}^f (t_i) &= M _{i-1} [\mathbf{x} ^a (t_{i-1} ) ]\\
+\mathbf{P}^f (t_i ) &= L_{i-1} \mathbf{P}^a (t_{i-1} ) L^T _{i-1} + \mathbf{Q} (t_{i-1} ) \end{align*}
+where $\mathbf{Q}$ is the noise covariance matrix (model error).
+In the experiments here, $\mathbf{Q} = 0$ since our model is perfect.
+In NWP, $\mathbf{Q}$ must be approximated, e.g. using statistical moments on the analysis increments \cite{danforth2007estimating,li2009accounting}.
+The analysis step is then written as (for $H$ the observation operator):
+\begin{align} \mathbf{x}^a (t_i ) &= \mathbf{x}^f (t_i) + \mathbf{K}_i \mathbf{d}_i\\
+\mathbf{P}^a (t_i) &= (\mathbf{I} - \mathbf{K}_i \mathbf{H}_i )\mathbf{P}^f (t_i) \end{align}
+where
+\[ \mathbf{d}_i = \mathbf{y}_i^o - \mathbf{H}[x^f (t_i) ] \]
+is the innovation. The Kalman gain matrix is computed to minimize the analysis error covariance $P^a _i$ as
+\[ \mathbf{K}_i = \mathbf{P}^f (t_i) \mathbf{H}_i ^T [ \mathbf{R}_i + \mathbf{H}_i \mathbf{P}^f (t_i) \mathbf{H}^T ] ^{-1} \]
+where $\mathbf{R}_i$ is the observation error covariance.
+Since we are making observations of the truth with known standard deviation $\mathbf{\epsilon}$, the observational error covariance matrix $\mathbf{R}$ is a diagonal matrix with the standard deviatoin $\epsilon$ along the diagonal.
+This information is an additional assumption, we could however not use this information and simply sample $\epsilon$ as a part of the experiment.
+The most difficult, and most computationally expensive, part of the EKF is deriving and integrating the TLM.
+Here we use a differentiated Runge-Kutta scheme of 2-nd order to accurately integrate the TLM.
+For more details on this implemenation, see Reagan \cite{reagan2013}.
+
 \subsection{Genetic Algorithm}
-\indent 
+
+\subsection{Experiment}
+
+For both of the systems, we study the performance of our parameter estimation scheme under varying observational noise, observational density, observational frequency, nonlinearity and dimension.
+This amounts to 4 (Lorenz 63) and 5 (Lorenz 96) dimensions of the experiment, and we outline the specific choices for each experiment in Table 1. %%\ref{table:experimentParms}.
+
   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{thebibliography}{9}
+
+        \bibitem{achtelik2009visual}
+        Achtelik, M., T.~Zhang, K.~Kuhnlenz, and M.~Buss (2009).
+        \newblock Visual tracking and control of a quadcopter using a stereo camera
+          system and inertial sensors.
+        \newblock In {\em Mechatronics and Automation, 2009. ICMA 2009. International
+          Conference on}, pp.\	2863--2869. IEEE.
+
+        \bibitem{siouris1997tracking}
+        Siouris, G.~M., G.~Chen, and J.~Wang (1997).
+        \newblock Tracking an incoming ballistic missile using an extended interval
+          kalman filter.
+        \newblock {\em Aerospace and Electronic Systems, IEEE Transactions on\/}~{\em
+          33\/}(1), 232--240.
+
+        \bibitem{talagrand1997assimilation}
+        Talagrand, O. (1997).
+        \newblock Assimilation of observations, an introduction.
+        \newblock {\em JOURNAL-METEOROLOGICAL SOCIETY OF JAAN SERIES 2\/}~{\em 75},
+          81--99.
+
+        \bibitem{kalman1961new}
+        Kalman, R. and R.~Bucy (1961).
+        \newblock New results in linear prediction and filtering theory.
+        \newblock {\em Trans. AMSE J. Basic Eng. D\/}~{\em 83}, 95--108.
+
+        \bibitem{savely1972}
+        Savely, R., B.~Cockrell, , and S.~Pines (1972).
+        \newblock Apollo experience report - onboard navigational and alignment
+          software.
+        \newblock {\em Technical Report\/}.
+
+        \bibitem{kalnay2003}
+        Kalnay, E. (2003).
+        \newblock {\em Atmospheric modeling, data assimilation, and predictability}.
+        \newblock Cambridge university press.
+
+        \bibitem{danforth2007estimating}
+        Danforth, C.~M., E.~Kalnay, and T.~Miyoshi (2007).
+        \newblock Estimating and correcting global weather model error.
+        \newblock {\em Monthly weather review\/}~{\em 135\/}(2), 281--299.
+
+        \bibitem{li2009accounting}
+        Li, H., E.~Kalnay, T.~Miyoshi, and C.~M. Danforth (2009).
+        \newblock Accounting for model errors in ensemble data assimilation.
+        \newblock {\em Monthly Weather Review\/}~{\em 137\/}(10), 3407--3419.
+
+        \bibitem{reagan2013}
+        Reagan, A (2013).
+        \newblock Predicting Flow Reversals in a Computational Fluid Dyanmics Simulated Thermosyphon Using Data Assimilatoin.
+        \newblock University of Vermont Master's Thesis, {\em arXiv:1312.2142 [math.DS]}.
+
+        \bibitem{kalnay20074}
+        Kalnay, E., H.~Li, T.~Miyoshi, S.-C. YANG, and J.~BALLABRERA-POY (2007).
+        \newblock 4-d-var or ensemble kalman filter?
+        \newblock {\em Tellus A\/}~{\em 59\/}(5), 758--773.
+
+        \bibitem{lorenz1963}
+        Lorenz, E.~N. (1963).
+        \newblock Deterministic nonperiodic flow.
+        \newblock {\em Journal of the atmospheric sciences\/}~{\em 20\/}(2), 130--141.
+
+        \bibitem{saltzman1962finite}
+        Saltzman, B. (1962).
+        \newblock Finite amplitude free convection as an initial value problem-i.
+        \newblock {\em Journal of the Atmospheric Sciences\/}~{\em 19\/}(4), 329--341.
+
 	\bibitem{kerr}
 	R. A. Kerr. \emph{Weather Forecasts Slowly Clearing Up}. Science 9 November 2012: Vol. 338 no. 6108 pp. 734-737 DOI: 10.1126/science.338.6108.734
+
         \bibitem{farmer}
                 Farmer, J. D., and J. J. Sidorowich. \emph{Predicting Chaotic Time Series.} Phys. Rev. Lett. 59(8) (1987): 845-848.
+
 	\bibitem{orrell3}

### 2013-12-11 project003/UVM/DA-code/runDAParamFitAll.sh
finished running for L63

@@ -23,7 +23,7 @@ cd /users/a/r/areagan/class/2013/MCRN501/project003/UVM/DA-code
 export NUMRUNS=100
 export RUNTIME=200
 ## part of experiment
-export RHO=22
+export RHO=35
 ## for normal error
 ERRORDIST="normal"
 export ERRORDIST
@@ -39,7 +39,13 @@ do
       for OBSVAR in 1 3
       do
 	export OBSVAR
-	qsub -q shortq -V run.qsub
+	FILENAME="data/${ERRORDIST}_${OBSERROR}_${NUMRUNS}_${RUNTIME}_${SUBSAMPLEALPHA}_${RHO}_${OBSVAR}_${EXPCOUNT}_forecastEnds.csv"
+	echo $FILENAME
+	if [ -f $FILENAME ]; then
+	  echo "already done"
+	else
+	  qsub -q workq -V run.qsub
+	fi
       done
     done
   done
@@ -59,8 +65,31 @@ do
       for OBSVAR in 1 3
       do
 	export OBSVAR
-	qsub -q shortq -V run.qsub
+	FILENAME="data/${ERRORDIST}_${OBSERROR}_${NUMRUNS}_${RUNTIME}_${SUBSAMPLEALPHA}_${RHO}_${OBSVAR}_${EXPCOUNT}_forecastEnds.csv"
+	echo $FILENAME
+	if [ -f $FILENAME ]; then
+	  echo "already done"
+	else
+	  qsub -q workq -V run.qsub
+	fi
       done
     done
   done
 done
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+

### 2013-12-11 project003/UVM/DA-code/run_filter.m

@@ -31,6 +31,9 @@ obsVarLen = checkenv('OBSVAR',3);
 obsVar = 1:obsVarLen;
 expCount = checkenv('EXPCOUNT',1);
 
+fprintf('save file is\n');
+fprintf('data/%s_%g_%g_%g_%g_%g_%g_%g_forecastEnds.csv\n',obs_error_type,obs_error_std,numRuns,runTime,windowAlpha,rho,obsVarLen,expCount);
+
 %% set things based off of this
 windowLen = windowAlpha*tStep;
 modelname = 'lorenz63';
@@ -95,7 +98,7 @@ allAnalyses = allForecasts;
 
 %% now run a lot of them
 for i=1:numRuns
-  disp(i);
+  fprintf('on numRun %g\n',i);
   [tmp_EKF_f_vec,tmp_EKF_a_vec] = modelDAinterface(@lorenz63_paramEst,'EKF',obs_pert,0:windowLen:runTime,H,R,windowLen,0,0);
   allForecasts(:,i+1) = tmp_EKF_f_vec(:,end);
   allAnalyses(:,i+1) = tmp_EKF_a_vec(:,end);
@@ -105,6 +108,9 @@ end
 %% forecasts
 csvwrite(sprintf('data/%s_%g_%g_%g_%g_%g_%g_%g_forecastEnds.csv',obs_error_type,obs_error_std,numRuns,runTime,windowAlpha,rho,obsVarLen,expCount),allForecasts);
 csvwrite(sprintf('data/%s_%g_%g_%g_%g_%g_%g_%g_analysisEnds.csv',obs_error_type,obs_error_std,numRuns,runTime,windowAlpha,rho,obsVarLen,expCount),allAnalyses);
+
+fprintf('success\n');
+
 
 
 

### 2013-12-11 project003/UVM/DA-code/EKF.m
other files needed to run on VACC

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/filters/EKF.m

### 2013-12-11 project003/UVM/DA-code/checkenv.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/checkenv.m

### 2013-12-11 project003/UVM/DA-code/lorenz63.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/lorenz63.m

### 2013-12-11 project003/UVM/DA-code/modelDAinterface.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/modelDAinterface.m

### 2013-12-11 project003/UVM/DA-code/rk2.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/rk2.m

### 2013-12-11 project003/UVM/DA-code/rk2prime.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/rk2prime.m

### 2013-12-11 project003/UVM/DA-code/runDAParamFitAll-wRho.sh

@@ -1,0 +1,98 @@
+#!/bin/bash
+
+# need these for compiled matlabs jobs
+export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/gpfs1/arch/x86_64/matlab2010a/bin/glnxa64:/gpfs1/arch/x86_64/matlab2010a/runtime/glnxa64:/gpfs1/arch/x86_64/matlab2010a/sys/os/glnxa64:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/native_threads:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/server:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64/client:/gpfs1/arch/x86_64/matlab2010a/sys/java/jre/glnxa64/jre/lib/amd64
+export XAPPLRESDIR=/gpfs1/arch/x86_64/matlab2010a/X11/app-defaults
+
+cd /users/a/r/areagan/class/2013/MCRN501/project003/UVM/DA-code
+
+# ## for a single test (always find something wrong)
+# EXPCOUNT=1 ## will run through 5 experiments
+# ## tunable parameters
+# NUMRUNS=100
+# RUNTIME=200
+# ## part of experiment
+# OBSERROR=0.05
+# ERRORDIST="normal" #"uniform"
+# SUBSAMPLEALPHA=1
+# RHO=28
+# OBSVAR=3
+# qsub -v NUMRUNS,RUNTIME,OBSERROR,ERRORDIST,SUBSAMPLEALPHA,RHO,OBSVAR,EXPCOUNT run.qsub
+
+## tunable parameters
+export NUMRUNS=100
+export RUNTIME=200
+## part of experiment
+for RHO in 22 28 35
+do
+  export RHO
+  ## for normal error
+  ERRORDIST="normal"
+  export ERRORDIST
+  for EXPCOUNT in {1..5} ## will run through 5 experiments
+  do
+    export EXPCOUNT
+    for OBSERROR in 0 0.01 0.05 0.1 0.25 0.5 1 2
+    do
+      export OBSERROR
+      for SUBSAMPLEALPHA in 1 5 25 50
+      do
+        export SUBSAMPLEALPHA
+        for OBSVAR in 1 3
+        do
+  	export OBSVAR
[…]

### 2013-12-11 project003/UVM/DA-code/run_filter_main.c

@@ -1,0 +1,130 @@
+/*
+ * MATLAB Compiler: 4.13 (R2010a)
+ * Date: Sun Dec  1 23:19:21 2013
+ * Arguments: "-B" "macro_default" "-m" "-W" "main" "-T" "link:exe"
+ * "run_filter.m" 
+ */
+#include <stdio.h>
+#include "mclmcrrt.h"
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+extern mclComponentData __MCC_run_filter_component_data;
+
+#ifdef __cplusplus
+}
+#endif
+
+static HMCRINSTANCE _mcr_inst = NULL;
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+static int mclDefaultPrintHandler(const char *s)
+{
+  return mclWrite(1 /* stdout */, s, sizeof(char)*strlen(s));
+}
+
+#ifdef __cplusplus
+} /* End extern "C" block */
+#endif
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+static int mclDefaultErrorHandler(const char *s)
+{
+  int written = 0;
+  size_t len = 0;
+  len = strlen(s);
+  written = mclWrite(2 /* stderr */, s, sizeof(char)*len);
+  if (len > 0 && s[ len-1 ] != '\n')
+    written += mclWrite(2 /* stderr */, "\n", sizeof(char));
+  return written;
+}
+
+#ifdef __cplusplus
+} /* End extern "C" block */
+#endif
+
+#ifndef LIB_run_filter_C_API
+#define LIB_run_filter_C_API /* No special import/export declaration */
+#endif
+
+LIB_run_filter_C_API 
+bool MW_CALL_CONV run_filterInitializeWithHandlers(
+    mclOutputHandlerFcn error_handler,
+    mclOutputHandlerFcn print_handler)
+{
+    int bResult = 0;
+  if (_mcr_inst != NULL)
+    return true;
+  if (!mclmcrInitialize())
+    return false;
+    {
+        mclCtfStream ctfStream = 
+            mclGetEmbeddedCtfStream((void *)(run_filterInitializeWithHandlers), 
+                                    33010);
+        if (ctfStream) {
+            bResult = mclInitializeComponentInstanceEmbedded(   &_mcr_inst,
+                                                                
+                                                     &__MCC_run_filter_component_data,
+                                                                true, 
+                                                                NoObjectType, 
+                                                                ExeTarget,
[…]

### 2013-12-11 project003/UVM/DA-code/run_filter_mcc_component_data.c

@@ -1,0 +1,150 @@
+/*
+ * MATLAB Compiler: 4.13 (R2010a)
+ * Date: Sun Dec  1 23:19:21 2013
+ * Arguments: "-B" "macro_default" "-m" "-W" "main" "-T" "link:exe"
+ * "run_filter.m" 
+ */
+
+#include "mclmcrrt.h"
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+const unsigned char __MCC_run_filter_session_key[] = {
+    'B', '6', 'D', '1', '7', 'E', 'E', '4', '8', '1', 'D', '2', '0', 'F', 'C',
+    '3', '3', 'E', 'D', '0', '8', 'B', '7', '0', '3', 'F', '2', '6', '5', '8',
+    '9', '0', '9', '8', '3', '9', '6', 'E', '2', '8', '6', 'C', '5', 'D', '8',
+    '5', '2', 'B', '3', '8', 'C', '0', '1', 'C', '6', '4', '7', 'F', 'B', 'A',
+    'D', '3', 'E', '8', 'A', 'B', '6', '1', '4', 'D', '4', 'E', '3', '6', '1',
+    'E', '1', 'E', '6', '4', 'A', 'B', '8', 'A', '8', '7', '3', 'C', 'D', 'F',
+    '9', '6', '1', '9', 'C', '1', 'E', 'E', '1', '5', '7', 'B', '3', '2', 'F',
+    'C', '2', '9', 'F', '6', '4', '8', '5', '1', 'F', '3', '0', '0', 'B', 'C',
+    '3', '7', 'E', '9', '3', '0', '6', '8', 'E', '2', '8', 'F', 'E', '5', '3',
+    '3', 'E', 'C', '4', '7', 'E', 'E', '8', '9', 'C', '6', '7', '8', '6', '2',
+    '9', 'F', 'E', '3', '2', 'C', 'D', '1', '8', '3', 'A', '2', 'D', '6', '9',
+    'D', '1', '9', '3', '2', 'D', '8', 'A', 'F', 'E', '0', 'A', 'C', 'C', 'F',
+    '3', 'E', '1', '8', 'A', '9', '2', '8', 'F', '8', '3', '5', '9', 'B', '7',
+    '9', '9', 'A', 'F', '2', '4', '8', 'E', 'C', '3', '6', 'A', '9', '9', 'F',
+    'F', '0', 'B', 'E', '1', 'A', 'A', '0', '5', '0', '4', 'D', 'B', 'D', '3',
[…]

### 2013-12-11 project003/UVM/DA-code/run_run_filter.sh

@@ -1,0 +1,33 @@
+#!/bin/sh
+# script for execution of deployed applications
+#
+# Sets up the MCR environment for the current $ARCH and executes 
+# the specified command.
+#
+exe_name=$0
+exe_dir=`dirname "$0"`
+echo "------------------------------------------"
+if [ "x$1" = "x" ]; then
+  echo Usage:
+  echo    $0 \<deployedMCRroot\> args
+else
+  echo Setting up environment variables
+  MCRROOT="$1"
+  echo ---
+  LD_LIBRARY_PATH=.:${MCRROOT}/runtime/glnxa64 ;
+  LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRROOT}/bin/glnxa64 ;
+  LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRROOT}/sys/os/glnxa64;
+	MCRJRE=${MCRROOT}/sys/java/jre/glnxa64/jre/lib/amd64 ;
+	LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRJRE}/native_threads ; 
+	LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRJRE}/server ;
+	LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRJRE}/client ;
+	LD_LIBRARY_PATH=${LD_LIBRARY_PATH}:${MCRJRE} ;  
+  XAPPLRESDIR=${MCRROOT}/X11/app-defaults ;
+  export LD_LIBRARY_PATH;
+  export XAPPLRESDIR;
+  echo LD_LIBRARY_PATH is ${LD_LIBRARY_PATH};
+  shift 1
+  "${exe_dir}"/run_filter $*
+fi
+exit
+

### 2013-12-11 project003/UVM/DA-code/lorenz96_paramEst.m
setting up L96

@@ -1,0 +1,154 @@
+classdef lorenz96_paramEst<handle
+	%lorenz96: class for a lorenz96 run
+	%   Goal is to implement a class for which MATLAB can
+	%   control and run the lorenz96 model
+	%
+	%   USAGE:
+	%       -initialize the model with a climatological IC
+	%        ens1 = lorenz96();
+	%        ens1.init();
+	%        OR with a name
+	%        ens1.init('name')
+	%        OR initialize with given IC (and name)
+	%        ens1.init('name',[10.4;11.7;18.5]);
+	%       -set the IC for the run to something specific
+	%        ens1.x = [10.4;11.7;18.5];
+	%       -run this case
+	%        ens1.run();
+	%       -run more into the future
+	%        ens1.run();
+	
+	properties
+		%basics
+		x
+		tstep = .01;
+		dim = 6;
+		time = 0;
+		window = 20;
+		
+		params = {};
+		%for the TLM
+		%this code only works for rk2prime...
+		TLMmethod = 'rk2prime'; % 'kam','explicit'
+		p_f
+		DIR
+	end % properties
+	methods
+		function self = lorenz96(varargin)
+			%intialize the class
+		end %constructor
+		function init(self,varargin)
+			self.x = 10*randn(self.dim,1);
+			self.x(4:6) = 100*eye(3)*rand(3,1); %3*randn(3,1)+[8/3;10;28]; %[10 0 0;0 20 0; 0 0 40]
+			%fprintf('initializing...');
+			
+			% ignore the first argument, if there is one
+			if nargin > 2
+				% setting given IC
+				self.x = varargin{2};
+			else
+				% generating climatological IC
+				tmpwindow = self.window; self.window = randi(100,1);
+				tmptime = self.time; self.time = 0;
+				self.x(1:3) = [-12.6480;-13.9758;30.9758]+.1*randn(3,1);
+				self.run();
+				self.window = tmpwindow;
+				self.time = tmptime;
+			end
+			%fprintf('done\n');
+		end %init
+		function run(self,varargin)
+			% run the case
+			%fprintf('running...');
+			
[…]

### 2013-12-11 project003/UVM/DA-code/lorenz96_paramEst.m
included the L96 model in the class, modified output column vector, using tail of vector for four parameters (in and out)

@@ -21,15 +21,15 @@ classdef lorenz96_paramEst<handle
 	properties
 		%basics
 		x
-		tstep = .01;
-		dim = 6;
+		tstep = .001;
+		dim = 20;
 		time = 0;
 		window = 20;
 		
 		params = {};
 		%for the TLM
 		%this code only works for rk2prime...
-		TLMmethod = 'rk2prime'; % 'kam','explicit'
+		TLMmethod = 'rk4prime'; % 'kam','explicit'
 		p_f
 		DIR
 	end % properties
@@ -61,7 +61,7 @@ classdef lorenz96_paramEst<handle
 			% run the case
 			%fprintf('running...');
 			
-			[~,tmp_x_f] = rk2(@lorenz96_model,self.params,self.time+[0,self.window],self.x,self.tstep);
+			[~,tmp_x_f] = rk4(@lorenz96_model,self.params,self.time+[0,self.window],self.x,self.tstep);
 			self.x = tmp_x_f(end,:)';
 			self.time = self.time+self.window;
 			%fprintf('done\n');
@@ -77,22 +77,50 @@ classdef lorenz96_paramEst<handle
 	end % methods
 end % classdef
 
-function [xprime] = lorenz96_model(~,x,~)
-% the Lorenz '96 system, as a function
-%
-% INPUT
-%   t  - time, scalar
-%   y  - column vector solution
+function dstate=lorenz96_model(~,state,~)
+global I;global J;
+h=state(end-3);c=state(end-2);b=state(end-1);F=state(end);
+state=reshape(state(1:end-4),I,J+1);
+IJ=I*J;
 
-b = x(4); s = x(5); r = x(6);
-xprime = [s*(x(2)-x(1));r*x(1)-x(2)-x(1)*x(3);x(1)*x(2)-b*x(3);0;0;0];
+X=state(:,1);
 
+state(:,1)=X(mod(-1:I-2,I)+1).*(X(mod(1:I,I)+1)-X(mod(-2:I-3,I)+1))-X+F*ones(I,1)-...
+    (h*c/b)*sum(state(:,2:end),2);
+
+IJ=I*J;
+Ys=zeros(1,IJ);
+Xs=zeros(1,IJ);
+%Ys is the outer ring of fast oscillators
+for i=1:I
[…]

### 2013-12-11 project003/UVM/DA-code/lorenz96_paramEst.m
incorporating the jac. big question here, are we actually going to predict the fast oscillators?

@@ -22,14 +22,15 @@ classdef lorenz96_paramEst<handle
 		%basics
 		x
 		tstep = .001;
-		dim = 20;
+        I = 10; % slow oscillators
+        J = 4; % fast guys per slow
+		dim = I*(J+1)+4; % add the parameters h,b,c,F
 		time = 0;
-		window = 20;
+		window = 1;
 		
 		params = {};
 		%for the TLM
-		%this code only works for rk2prime...
-		TLMmethod = 'rk4prime'; % 'kam','explicit'
+		TLMmethod = 'rk4prime';
 		p_f
 		DIR
 	end % properties
@@ -38,9 +39,11 @@ classdef lorenz96_paramEst<handle
 			%intialize the class
 		end %constructor
 		function init(self,varargin)
-			self.x = 10*randn(self.dim,1);
-			self.x(4:6) = 100*eye(3)*rand(3,1); %3*randn(3,1)+[8/3;10;28]; %[10 0 0;0 20 0; 0 0 40]
-			%fprintf('initializing...');
+            self.dim = self.I*(self.J+1)+4;
+            self.params = {self.I,self.J};
+            
+			self.x = rand(self.dim,1);
+			self.x(end-3:end) = 100*rand(4,1);
 			
 			% ignore the first argument, if there is one
 			if nargin > 2
@@ -48,14 +51,12 @@ classdef lorenz96_paramEst<handle
 				self.x = varargin{2};
 			else
 				% generating climatological IC
-				tmpwindow = self.window; self.window = randi(100,1);
+				tmpwindow = self.window; self.window = randi(10,1);
 				tmptime = self.time; self.time = 0;
-				self.x(1:3) = [-12.6480;-13.9758;30.9758]+.1*randn(3,1);
 				self.run();
 				self.window = tmpwindow;
 				self.time = tmptime;
-			end
-			%fprintf('done\n');
+            end
 		end %init
 		function run(self,varargin)
 			% run the case
@@ -77,8 +78,9 @@ classdef lorenz96_paramEst<handle
 	end % methods
 end % classdef
 
-function dstate=lorenz96_model(~,state,~)
-global I;global J;
+function dstate=lorenz96_model(~,state,params)
+I = params{1};
[…]

### 2013-12-12 project003/UVM/DA-code/myMod.m
code added

@@ -1,0 +1,10 @@
+function i=myMod(i,s)
+% disp([i,s])
+if i<1
+    i=s+i;
+elseif i>s
+    i=myMod(i-s,s);
+end;
+% disp(i)
+% disp('----------')
+end

### 2013-12-12 project003/UVM/DA-code/rk4.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/rk4.m

### 2013-12-12 project003/UVM/DA-code/rk4prime.m

@@ -1,0 +1,1 @@
+/users/a/r/areagan/work/2013/data-assimilation/src/models/rk4prime.m

### 2013-12-13 project003/UVM/finalPaper/paper.tex
one more line
added Morgan stuff to paper

@@
 
+\indent The area of genetic algorithms (GAs) is a prominent area of research \cite{GA1,GA2} with applications in several areas of research, including bankruptcy modeling \cite{Shin2002321}, calibrating water runoff models \cite{WRR}, and spectral data analysis \cite{CEM}.
+One key feature of GAs is that no knowledge of the model being fitted is required.
+To demonstrate the availability and robustness of this algorithm, we proceed by utilizing Matlab's built-in GA function, called ``{\it ga.m}", with essentially no alterations from the default options (i.e.
+the defaults for ``{\it gaoptimset.m}").
+The only assumption of note is that we assume parameters are positive real-values.
+Like other evolutionary algorithms, GA is applicable whenever a problem can be phrased in the biological evolution paradigm.
+The major hallmarks of this paradigm include identifying a population of genes and subjecting that population to crossover, genetic mutation, and selection pressure.
+The ``indivi!
+ duals" in the population of genes for our experiments will be real-valued vectors where each entry in a vector represents a parameter choice, and vectors are of the same length as the number of parameters being recovered.
+\\
+\indent Selection pressure is imposed on our population through a fitness-evaluation function that evaluates the ``fitness" of an individual by the the root-mean square error of a model integration with the parameter choices encoded in the individual vector who's fitness is being evaluated.
+The root-mean square error is calculated at each time $t$ that we have observation data.
+Notice that lower fitness is better in this context.
+Furthermore, we are attempting to recover parameters for chaotic systems; thus, there may exist brief intervals where the integration fits the observed data simply by chance.
+This has to do with the unpredictability of chaotic systems  over long integration times, and the bounded nature of the chaotic attractors in this study.
+We address this concern by restarting our integration at unit model-time intervals based on the observation data.\\
+\indent We use stochastic uniform selection to select two individuals at a time based on their fitness to undergo single-point crossover.
+Single-point crossover begins by randomly selecting a  vector index.
+The two ``parents" undergoing crossover are replaced in the population by two ``children".
+The first child is conceived by taking the indices from the first parent up to the selected index.
+The remaining indices are filled from the second parent.
+The second child is created following the same process where the parents' roles are switched.
+This process allows the population to converge on good parameter choices since a good parameter choice will yield favorable fitness scores.
+Fit individuals will have an increased chance of being selected to pass on genetic material to the next generation.\\
+\indent Genetic mutation is the mechanisms that allows the population to further explore parameter space.
+When an individual is subject to mutation, a vector index is selected randomly and a randomly selected small real-value is added to or subtracted from the value currently stored at that index of the individual.
+Large population sizes along with a high mutation rate (i.e.
+the probability of being subject to mutation) encourage robust sampling of parameter space.\\
+\indent The above mentioned mechanisms are highly tunable in that mutation, crossover, fitness evaluation, and selection can all be achieved through a variety of different methods.
+The key is to pick these methods to fit the problem you are addressing.
+The GA is initialized by randomly generating a population of real-valued vectors of the same length as the number of parameters we are trying to recover.
+Each individual is then assessed for fitness.
+Now individuals are selected for reproduction based on their fitness and other individuals are subject to genetic mutation.
+The two fittest individuals from the population are allowed to remain unchanged.
+This process yields the population for the next generation of individuals.
+We iterate this procedure for a prescribed number of generations, or until the population-wide fitness stops changing with successive generations.
+Figure 2 shows examples for some of these mechanisms along with a snap shot from a GA experiment attempting to recover parameters for L63.
+
 \subsection{Experiment}
@@
 This amounts to 4 (Lorenz 63) and 5 (Lorenz 96) dimensions of the experiment, and we outline the specific choices for each experiment in Table 1. %%\ref{table:experimentParms}.
+The experiments were chosen to mimic realisitic conditions under which simple models are fit to data.
 
@@
 \section{Discussion}
+
+\clearpage
+\begin{center}
+	\includegraphics[scale=.34]{GA_exmaple.png}\\
+	{\flushleft Example GA experiment:}\\
+	$\begin{array}{cc}
+		\fbox{\includegraphics[width=.27\columnwidth]{sample.pdf}}&
+		\hspace{-.3cm}\fbox{\includegraphics[width=.27\columnwidth]{sample2.pdf}}\\
+		\fbox{\includegraphics[width=.27\columnwidth]{sample3.pdf}}&		
+		\hspace{-.3cm}\fbox{\includegraphics[width=.27\columnwidth]{sample4.pdf}}				
+	\end{array}$
+\end{center}
+{\bf\small Figure 2.
+(Top) A cartoon demonstrating the flow of control for GA execution.
+The population of random real-valued vectors is initialized, and each individual has its fitness assessed.
+With some probability individuals are selected for single-point crossover or mutation.
+The children from these processes are collected into the new population.
+This process is iterated for a prescribed number of generations or until iteration fails to yield improvement.
+(Bottom) An example illustrating the improvement in parameter estimation made by the GA when attempting to recover the parameters ($\sigma=10,b=8/3,R=28$) over 100 generations.
+The observed data (truth) is in \textcolor{blue}{blue}, while the trajectory yielded from the best guess at the true parameters is provided in \textcolor{red}{red}.
+From left to right, we show the best solution after 10, 35, 52, and 100 generations.
+We see that at 100 generations, the best guess at the true parameters is reasonably close to th!
+ e correct answer ($\sigma=9.99711,b=2.66688,R=28.0004$)}\\
+\clearpage
 
@@
          D. S. Wilks, \emph{Effects of Stochastic Parametrizations in the Lorenz Õ96 System}, Quart. J. Roy. Meteo. Soc. 131 (2005) 389Ð407.
+
+	\bibitem{GA1}
+		J.
+Horn, N.
+Nafpliotis, and D.E.
+Goldberg.
+\emph{A niched Pareto genetic algorithm for multiobjective optimization}.
+Evolutionary Computation, 1994.
+IEEE World Congress on Computational Intelligence., Proceedings of the First IEEE Conference on (1994) 1: 82-87.
+doi:10.1109/ICEC.1994.350037
+%@INPROCEEDINGS{GA1, 
+%author={Horn, J. and Nafpliotis, N. and Goldberg, D.E.}, 
+%booktitle={Evolutionary Computation, 1994. IEEE World Congress on Computational Intelligence., Proceedings of the First IEEE Conference on}, 
+%title={A niched Pareto genetic algorithm for multiobjective optimization}, 
+%year={1994}, 
+%pages={82-87 vol.1}, 
+%keywords={genetic algorithms;operations research;optimisation;Niched Pareto GA;Pareto genetic algorithm;Pareto optimal set;genetic algorithm;multiobjective optimization;multiple objectives;niching pressure;optimization problems;Bioinformatics;Contracts;Costs;Genetic algorithms;Genomics;Internet;NASA;Pareto optimization;Sampling methods;Turning},
+%doi={10.1109/ICEC.1994.350037}}
+	\bibitem{GA2}
+		K.
+Deb, A.
+Pratap, S.
+Agarwal, and T.
+Meyarivan.
+\emph{A fast and elitist multiobjective genetic algorithm: NSGA-II}.
+Evolutionary Computation, IEEE Transactions on 6 (2): 182-197.
+doi:10.1109/4235.996017
+%@ARTICLE{GA1, 
+%author={Deb, K. and Pratap, A. and Agarwal, S. and Meyarivan, T.}, 
+%journal={Evolutionary Computation, IEEE Transactions on}, 
+%title={A fast and elitist multiobjective genetic algorithm: NSGA-II}, 
+%year={2002}, 
+%volume={6}, 
+%number={2}, 
+%pages={182-197}, 
+%keywords={Pareto distribution;computational complexity;constraint theory;convergence;genetic algorithms;operations research;simulation;sorting;NSGA-II;Nondominated Sorting Genetic Algorithm II;Pareto-archived evolution strategy;Pareto-optimal front;algorithm performance;computational complexity;constrained multi-objective problems;constraint handling;convergence;dominance definition;fast elitist multi-objective genetic algorithm;mating pool;multi-criterion decision making;multi-objective evolutionary algorithm;multi-objective optimization;nondominated sharing;nonlinear problem;objectives;parent/offspring population combination;population size;selection operator;simulation;solution fitness;solution spread;strength-Pareto evolutionary algorithm;Associate members;Computational complexity;Computational modeling;Constraint optimization;Decision making;Diversity reception;Evolutionary computation;Genetic algorithms;Sorting;Testing},
+%doi={10.1109/4235.996017}, 
+%ISSN={1089-778X}}
+
+	\bibitem{Shin2002321}
+		K-S Shin, Y-J Lee.
+\emph{A genetic algorithm application in bankruptcy prediction modeling}.
+Expert Systems w/ Applications (2002) 23 (3): 321-328.
+doi = "http://dx.doi.org/10.1016/S0957-4174(02)00051-9"
+%@article{Shin2002321,
+%title = "A genetic algorithm application in bankruptcy prediction modeling ",
+%journal = "Expert Systems with Applications ",
+%volume = "23",
+%number = "3",
+%pages = "321 - 328",
+%year = "2002",
+%note = "",
+%issn = "0957-4174",
+%doi = "http://dx.doi.org/10.1016/S0957-4174(02)00051-9",
+%url = "http://www.sciencedirect.com/science/article/pii/S0957417402000519",
+%author = "Kyung-Shik Shin and Yong-Joo Lee",
+%keywords = "Genetic algorithms",
+%keywords = "Rule extraction",
+%keywords = "Bankruptcy prediction "
+%}
+	\bibitem{WRR}
+		Q. J. Wang. \emph{The Genetic Algorithm and its Application to Calibrating Conceptual Rainfall-Runoff Models}. Water Resources Research 27 (9): 2467-2471. dpi:http://dx.doi.org/10.1029/91WR01305
+%@article {WRR,
+%author = {Wang, Q. J.},
+%title = {The Genetic Algorithm and Its Application to Calibrating Conceptual Rainfall-Runoff Models},
+%journal = {Water Resources Research},
+%volume = {27},
+%number = {9},
+%issn = {1944-7973},
+%url = {http://dx.doi.org/10.1029/91WR01305},
+%doi = {10.1029/91WR01305},
+%pages = {2467--2471},
+%year = {1991},
+%}
+
+\bibitem{CEM}
+R.
+Leardi.
+\emph{Application of genetic algorithm-PLS for feature selection in spectral data sets}.
+Journal of Chemometrics 14 (5-6): 643-655: DOI: 10.1002/1099-128X(200009/12)14:5/6<643::AID-CEM621>3.0.CO;2-E
+%@article {CEM,
+%author = {Leardi, Riccardo},
+%title = {Application of genetic algorithm窶撤LS for feature selection in spectral data sets},
+%journal = {Journal of Chemometrics},
+%volume = {14},
+%number = {5-6},
+%publisher = {John Wiley & Sons, Ltd.},
+%issn = {1099-128X},
+%url = {http://dx.doi.org/10.1002/1099-128X(200009/12)14:5/6<643::AID-CEM621>3.0.CO;2-E},
+%doi = {10.1002/1099-128X(200009/12)14:5/6<643::AID-CEM621>3.0.CO;2-E},
+%pages = {643--655},
+%keywords = {genetic algorithms, feature selection, PLS regression, spectral data},
+%year = {2000},
+%}
+
 \end{thebibliography}
 \end{multicols}
+
+\section{Appendix}
+\subsection{Experimental Results}
+\input{tableOutput.txt}
+
 \end{document}

### 2013-12-14 project003/UVM/finalPaper/paper.tex
my changes to final paper

@@
 From left to right, we show the best solution after 10, 35, 52, and 100 generations.
-We see that at 100 generations, the best guess at the true parameters is reasonably close to th!
- e correct answer ($\sigma=9.99711,b=2.66688,R=28.0004$)}\\
+We see that at 100 generations, the best guess at the true parameters is reasonably close to the correct answer ($\sigma=9.99711,b=2.66688,R=28.0004$)}\\
+
+\clearpage
+\begin{center}
+	\includegraphics[scale=1.00]{DA-figure.pdf}\\
+\end{center}
+{\bf\small  Example DA experiment: showing the $x_1$ variable of the Lorenz 63 system, with a observation subsampling of 25.
+With observational error, the EKF predicts the data well, frequently making analyses that are closer to the truth than the observations.}\\
 \clearpage
@@
 \subsection{Experimental Results}
+
 \input{tableOutput.txt}
 
+\input{../DA-code/allTables.tex}
+
 \end{document}

### 2013-12-14 project003/UVM/DA-code/makeTables.py

@@ -1,0 +1,327 @@
+#!/opt/local/bin/python
+
+## tunable parameters
+numRuns=100
+runTime=200
+
+def squared(tmpFloat):
+  value = tmpFloat**2
+  return value
+
+from numpy import sqrt,array
+
+obsErrorDict = dict()
+obsErrorDict['normal']=[0,0.01,0.05,0.1,0.25,0.5,1,2]
+obsErrorDict['uniform']=[0,0.5,2,4,6,8,10]
+
+allResults = dict()
+allResultErrors = dict()
+print 'loading data'
+
+for errorDist in ['normal','uniform']:
+  allResults[errorDist] = dict()
+  allResultErrors[errorDist] = dict()
+  for rho in [22,28,35]:
+    allResults[errorDist][rho] = dict()
+    allResultErrors[errorDist][rho] = dict()
+    for numObsVar in [1,3]:
+      allResults[errorDist][rho][numObsVar] = dict()
+      allResultErrors[errorDist][rho][numObsVar] = dict()
+      for obsError in obsErrorDict[errorDist]:
+        allResults[errorDist][rho][numObsVar][obsError] = dict()
+        allResultErrors[errorDist][rho][numObsVar][obsError] = dict()
+        for subSampleAlpha in [1,5,25,50]:
+	  allResults[errorDist][rho][numObsVar][obsError][subSampleAlpha] = []
+          allResultErrors[errorDist][rho][numObsVar][obsError][subSampleAlpha] = []
+	  for expCount in range(1,6):
+	    ## "data/${ERRORDIST}_${OBSERROR}_${NUMRUNS}_${RUNTIME}_${SUBSAMPLEALPHA}_${RHO}_${OBSVAR}_${EXPCOUNT}_forecastEnds.csv"
+  	    datafName="data/{0}_{1:g}_{2}_{3}_{4}_{5}_{6}_{7}_forecastEnds.csv".format(errorDist,obsError,numRuns,runTime,subSampleAlpha,rho,numObsVar,expCount)
+	    ## try:
+	    f = open(datafName,'r')
+	    ## fileOpened = 1
+	    tmpResults = []
+	    for line in f:
+	      tmpResults.append(map(float,line.rstrip().split(',')))
+	    f.close()
+            ## except:
+	      ## fileOpened = 0
+	      ## print 'could not open file {}'.format(datafName)
+	    ## if fileOpened:
+	    resultsResults = []
+	      ## print tmpResults
+	    for i in range(1,len(tmpResults[0])):
[…]
